### 2020-02-01 paper/paper.tex
Added paper

\documentclass[11pt,a4paper]{article}
\usepackage{jheppub}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{graphicx}
\newcommand{\expval}[1]{\left\langle#1\right\rangle}
\newcommand{\dd}{\mathrm{d}}

\title{One-loop effective action for the average Ising magnetization on compact manifolds}
\author{Andrea Allais}
\emailAdd{allais.andrea@gmail.com}

\begin{document}

\maketitle

\section{Main results}

The average magnetization $\bar{\phi} = \frac{1}{V} \sum_{i} \sigma_i$ of the
Ising model is the simplest observable in Monte Carlo simulations. For large
system size, the probability distribution of $\bar{\phi}$ changes qualitatively
between the two phases of the model. In the symmetric phase, if the
correlation length $\xi$ is large in lattice units, but small compared to the
system size $R$, the distribution is well approximated by a unimodal Gaussian
centered at 0. In the broken symmetry phase, still for $1 \ll \xi \ll R$, the
distribution is well approximated by a bimodal superposition of two 
Gaussians centered at non-zero values $\pm \bar{\phi}_0$.

Near the critical point $1 \ll R \ll \xi$ the probability distribution is a
non-trivial function of $\bar{\phi}$, at least in two and three dimensions.  It
is not immediately clear if the same is true in four dimensions as well.  Since
the $\phi^4$ field theory that describes the Ising critical point in four
dimensions becomes weakly coupled in the infrared, one may expect the
probability distribution to be a Gaussian at the critical point as well.

To clarify this issue, we compute the one-loop effective action, or log
probability, for $\bar{\phi}$ in $4$ dimensions, and show that for
very large system size, the critical probability distribution is not Gaussian,
but rather of the form $p(\bar{\phi})\sim \exp(-a \bar{\phi}^4)$. We find that
corrections to this form vanish very slowly with increasing system size, like
$1/\log R$.

Thus we consider the $\phi^4$ field theory:

\begin{equation}
  S[\phi] = \int_{\mathcal{M}}\left(
    \frac{1}{2} \phi\left(\Delta + r_0\right)\phi +
    \frac{1}{2} \xi_0 \phi^2 \mathcal{R} +
    \frac{1}{24} u_0 \phi^4\right)
\end{equation}

on both the 4-torus $\mathcal{M} = S_1^4$ or the 4-sphere $\mathcal{M} = S_4$.
Here $\mathcal R$ denotes the scalar curvature. We define the effective action:

\begin{equation}
  e^{-S_{\mathrm{eff}}(\bar\phi)} = \int \left[\mathcal{D}\phi\right]
    \delta\left(\bar{\phi} - \frac{1}{V}\int_{\mathcal{M}}\phi\right)
    e^{-S[\phi]}\,,
\end{equation}

and we obtain:

\begin{equation}
\label{eq:effective_action}
\begin{split}
  \frac{S_{\mathrm{eff}}(\bar{\phi})}{V} = 
  &\frac{1}{2} \frac{\bar{\phi}^2}{R^2} \left(
    r R^2 + \eta + \frac{g}{6} f_1\left(r R^2 + \eta\right)
    + O\left(g^2\right)\right) + \\
  & \frac{2\pi^2}{9}g\bar{\phi}^4\left(
    1 - \frac{g}{2}f_2\left(r R^2 + \eta\right)
    + O\left(g^2\right)\right) + \\
  & \frac{8\pi^4}{81} g^3 \bar{\phi}^6 R^2\left(
    f_3\left(r R^2 + \eta\right)
    + O\left(g\right)\right) + O\left(g^4\bar{\phi}^8\right)\,,
\end{split}
\end{equation}

where $R$ is the radius of either manifold, and the coefficients
$(r,\,\eta,\,g)$ are the renormalized counterparts to $(r_0,\,\xi,\,u_0)$. An
explicit expression for the functions $f_1$, $f_2$ and $f_3$ is reported in the
appendix (\ref{eq:f_functions}), and they are plotted in
fig.~\ref{fig:f_functions}.

\begin{figure}
\begin{center}
\includegraphics{f_functions.png}
\end{center}
\caption{\label{fig:f_functions} The functions $f_i$}
\end{figure}

This is our central result, and we will now describe it at lenght, beginning
with its domain of validity and its dependence on the couplings on a manifold
of fixed radius $R$.

The coupling $\eta$ is the counterpart to the curvature coupling
$\xi_0$, and hence is exactly zero for the torus, whereas it is a free,
dimensionless parameter for the sphere.

The dimension-2 coupling $r$ controls the cross-over between the symmetric and
broken-symmetry phases. The expression (\ref{eq:effective_action}) is valid for
$r R^2 + \eta \gtrsim -1$. It becomes invalid for lower values of $r$ because
the condensation of higher modes of the laplacian invalidates the perturbative
expansion. This becomes manifest in a divergence of the functions $f_i$ as
their argument approaches -1 for the torus, or  -4 for the sphere.  On the
other hand, $f_1$, $f_2$, $f_3$ remain finite as their argument approaches
positive infinity, and hence for large $r$ the probability distribution of
$\bar{\phi}$ is a Gaussian:

\begin{equation}
  \frac{S_{\mathrm{eff}}(\bar{\phi})}{V} \sim \frac{1}{2} r \bar{\phi}^2
  \quad \text{for} \quad
  rR^2 \gg 1\,.
\end{equation}

The dimensionless coupling $g$ is the parameter of the perturbative expansion.
The expansion is valid for $g \ll 1$, provided $rR^2$ is sufficiently far from
the bound discussed above.

In general,
they depend on a renormalization scale $\mu$, and it is possible to derive an
expression for $S_{\mathrm{eff}}$ evaluated at a generic value of $\mu$. Here,
for simplicity we choose to evaluate (\ref{eq:effective_action}) at the scale
$\mu^2 = r + p^2$. This value of $\mu$ is optimal for the reliability of
perturbation theory, because it avoids the emergence of large logarithms.

  As is well
known, in $4 - \varepsilon$ dimensions it satisfies the renormalization group
equation:

\begin{equation}
  \mu \frac{\dd g}{\dd \mu} = -\varepsilon g + g^2\,.
\end{equation}

The Wilson-Fisher fixed point is attained at $g_{\star} = \varepsilon$, and
substituting $g \to \varepsilon$ in (\ref{eq:effective_action}) and dropping
higher order terms in $\varepsilon$ yields the effective action at the fixed
point to leading order in $\epsilon$.

The coupling $r$ satisfies the renormalization group equation:
\begin{equation}
  \mu \frac{\dd r}{\dd \mu} = \frac{g}{3}r
\end{equation}

$r$, $\eta$, and $g$ are renormalized couplings
that satisfy the renormalization group equations:

\begin{align}
  &\mu \frac{\dd g}{\dd \mu} = -\varepsilon g + g^2 \\
  &\mu \frac{\dd r}{\dd \mu} = \frac{g}{3}r \\
  &\mu \frac{\dd \eta}{\dd \mu} = \frac{g}{3}(\eta - 2)
\end{align}

The coupling $\eta$ is related to the coupling between the scalar and
the curvature, and hence it is exactly zero on the tourus. On the sphere, it
may seem that it can be set to a constant value $\eta = 2$ at all energy
scales, however this ceases to be true at higher orders in perturbation theory,
where the renormalization group equation becomes inhomogeneous.

While a general expression for the effective action at a generic scale $\mu$ is
available, for simplicity (\ref{eq:effective_action}) is evaluated at $\mu^2
= r + p^2$. This value of $\mu$ is optimal for the reliability of perturbation
theory, because it minimizes the effect of the so-called large logs.

An explicit expression for the functions $f_1$, $f_2$, $f_3$ is reported in the
appendix. As shown in fig.~\ref{fig:f_functions}, $f_i(x)$ is of order 1 in for
$x \gtrsim -1/2$, $f_1$ approaches a constant as $x \to \infty$, whereas
$f_2$ and $f_3$ approach 0. They all diverge as $x$ approaches $-1$.
\section{Appendix}

\begin{equation}
\label{eq:f_functions}
f_1 = 
\end{equation}
\end{document}

### 2020-03-08 paper/paper.tex
Paper checkpoint

@@
 \usepackage{graphicx}
+\usepackage{tikz}
 \newcommand{\expval}[1]{\left\langle#1\right\rangle}
 \newcommand{\dd}{\mathrm{d}}
+\renewcommand{\vec}[1]{\boldsymbol{#1}}
 
-\title{One-loop effective action for the average Ising magnetization on compact manifolds}
+\title{One-loop PDF of the average Ising magnetization in 4 dimensions.}
 \author{Andrea Allais}
@@
 system size, the probability distribution of $\bar{\phi}$ changes qualitatively
-between the two phases of the model. In the symmetric phase, if the
-correlation length $\xi$ is large in lattice units, but small compared to the
-system size $R$, the distribution is well approximated by a unimodal Gaussian
-centered at 0. In the broken symmetry phase, still for $1 \ll \xi \ll R$, the
-distribution is well approximated by a bimodal superposition of two 
-Gaussians centered at non-zero values $\pm \bar{\phi}_0$.
+between the two phases of the model. In the symmetric phase, if the correlation
+length $\xi$ is large in lattice units, but small compared to the system size
+$L$, the distribution is well approximated by a normal distribution centered at
+0. In the broken symmetry phase, still for $1 \ll \xi \ll L$, the distribution
+is well approximated by a bimodal superposition of two normals centered at
+non-zero values $\pm \bar{\phi}_0$.
 
-Near the critical point $1 \ll R \ll \xi$ the probability distribution is a
-non-trivial function of $\bar{\phi}$, at least in two and three dimensions.  It
-is not immediately clear if the same is true in four dimensions as well.  Since
-the $\phi^4$ field theory that describes the Ising critical point in four
-dimensions becomes weakly coupled in the infrared, one may expect the
-probability distribution to be a Gaussian at the critical point as well.
+Near the critical point $1 \ll L \ll \xi$ the probability distribution of
+$\bar{\phi}$ is a non-trivial function of $\bar{\phi}$, at least in two and
+three dimensions.  It is not immediately clear if the same is true in four
+dimensions as well.  Since the $\phi^4$ field theory that describes the Ising
+critical point in four dimensions becomes weakly coupled in the infrared, one
+may expect the probability distribution to be a single zero-centered normal at
+the critical point as well.
 
-To clarify this issue, we compute the one-loop effective action, or log
-probability, for $\bar{\phi}$ in $4$ dimensions, and show that for
-very large system size, the critical probability distribution is not Gaussian,
-but rather of the form $p(\bar{\phi})\sim \exp(-a \bar{\phi}^4)$. We find that
-corrections to this form vanish very slowly with increasing system size, like
-$1/\log R$.
+To clarify this issue, we compute the log pdf of $\bar{\phi}$ at one-loop in
+$4$ dimensions, and show that for very large system size, the critical
+probability distribution is not normal, but rather of the form
+$-\log p(\bar{\phi})\sim f(L) \bar{\phi}^4$. We find that corrections to this
+form vanish very slowly with increasing system size, like $1/\log L$.
 
-Thus we consider the $\phi^4$ field theory:
+\begin{figure}
+\begin{center}
+\includegraphics[scale=0.75]{binder_cumulant.png}
+\end{center}
+\caption{\label{fig:binder_cumulant} Blah Blah}
+\end{figure}
 
+Thus we consider the $\phi^4$ field theory with action:
 \begin{equation}
-  S[\phi] = \int_{\mathcal{M}}\left(
+  S[\phi] \equiv \int \dd^4 x\ \left(
     \frac{1}{2} \phi\left(\Delta + r_0\right)\phi +
-    \frac{1}{2} \xi_0 \phi^2 \mathcal{R} +
     \frac{1}{24} u_0 \phi^4\right)
 \end{equation}
-
-on both the 4-torus $\mathcal{M} = S_1^4$ or the 4-sphere $\mathcal{M} = S_4$.
-Here $\mathcal R$ denotes the scalar curvature. We define the effective action:
-
+defined on a 4-torus of radius $R$, i.e. $x_i\sim x_i + 2\pi R$. In this field
+theory context, the pdf of the average magnetization is defined as:
 \begin{equation}
-  e^{-S_{\mathrm{eff}}(\bar\phi)} = \int \left[\mathcal{D}\phi\right]
-    \delta\left(\bar{\phi} - \frac{1}{V}\int_{\mathcal{M}}\phi\right)
+    \label{eq:pdf_definition}
+    p(\bar\phi) \equiv \frac{1}{Z} \int \left[\mathcal{D}\phi\right]
+    \delta\left(\bar{\phi} - \frac{1}{V}\int\dd^4 x\ \phi\right)
     e^{-S[\phi]}\,,
 \end{equation}
-
-and we obtain:
-
+and at one loop it evaluates to:
 \begin{equation}
-\label{eq:effective_action}
+\label{eq:log_pdf}
 \begin{split}
-  \frac{S_{\mathrm{eff}}(\bar{\phi})}{V} = 
+    -\log p(\bar{\phi}) = \mathrm{const.} +  V\Bigg(
   &\frac{1}{2} \frac{\bar{\phi}^2}{R^2} \left(
-    r R^2 + \eta + \frac{g}{6} f_1\left(r R^2 + \eta\right)
+    r R^2 + \frac{g}{6} \left(f_1\left(r R^2\right) - 1\right)
     + O\left(g^2\right)\right) + \\
   & \frac{2\pi^2}{9}g\bar{\phi}^4\left(
-    1 - \frac{g}{2}f_2\left(r R^2 + \eta\right)
+    1 - \frac{g}{2}f_2\left(r R^2\right)
     + O\left(g^2\right)\right) + \\
   & \frac{8\pi^4}{81} g^3 \bar{\phi}^6 R^2\left(
-    f_3\left(r R^2 + \eta\right)
-    + O\left(g\right)\right) + O\left(g^4\bar{\phi}^8\right)\,,
+    f_3\left(r R^2\right)
+    + O\left(g\right)\right) + O\left(g^4\bar{\phi}^8\right) \Bigg)\,,
 \end{split}
 \end{equation}
-
-where $R$ is the radius of either manifold, and the coefficients
-$(r,\,\eta,\,g)$ are the renormalized counterparts to $(r_0,\,\xi,\,u_0)$. An
-explicit expression for the functions $f_1$, $f_2$ and $f_3$ is reported in the
-appendix (\ref{eq:f_functions}), and they are plotted in
+where the coefficients $r$ and $g$ are the renormalized counterparts to $r_0$
+and $u_0$. An explicit expression for the functions $f_1$, $f_2$ and $f_3$ is
+reported in the appendix (\ref{eq:f_functions}), and they are plotted in
 fig.~\ref{fig:f_functions}.
@@
 \begin{center}
-\includegraphics{f_functions.png}
+\includegraphics[scale=0.75]{f_functions.png}
 \end{center}
@@
 
-This is our central result, and we will now describe it at lenght, beginning
-with its domain of validity and its dependence on the couplings on a manifold
-of fixed radius $R$.
-
-The coupling $\eta$ is the counterpart to the curvature coupling
-$\xi_0$, and hence is exactly zero for the torus, whereas it is a free,
-dimensionless parameter for the sphere.
+Eq. (\ref{eq:log_pdf}) is our central result, and we now describe it
+at lenght. We first illustrate its domain of validity and its dependence on the
+renormalized couplings $r$, $g$ on a torus of fixed radius $R$, then describe
+its dependence on $R$ at fixed bare couplings $r_0$, $u_0$.
 
 The dimension-2 coupling $r$ controls the cross-over between the symmetric and
-broken-symmetry phases. The expression (\ref{eq:effective_action}) is valid for
-$r R^2 + \eta \gtrsim -1$. It becomes invalid for lower values of $r$ because
-the condensation of higher modes of the laplacian invalidates the perturbative
-expansion. This becomes manifest in a divergence of the functions $f_i$ as
-their argument approaches -1 for the torus, or  -4 for the sphere.  On the
-other hand, $f_1$, $f_2$, $f_3$ remain finite as their argument approaches
-positive infinity, and hence for large $r$ the probability distribution of
-$\bar{\phi}$ is a Gaussian:
-
+broken-symmetry phases. The expression (\ref{eq:log_pdf}) is valid for
+all positive values of $r$, and also for negative values, provided that $rR^2
+\gtrsim -1/2$.  For lower values of $r$, additional modes of the laplacian
+condense beside the zero-mode, and this invalidates the perturbative expansion.
+. This becomes manifest in a divergence of the functions $f_i$ as their
+argument approaches -1.  On the other hand, $f_1$, $f_2$, $f_3$ all go to zero
+as their argument approaches positive infinity, and hence for large $r$ the
+probability distribution of $\bar{\phi}$ is a zero-centered normal:
 \begin{equation}
-  \frac{S_{\mathrm{eff}}(\bar{\phi})}{V} \sim \frac{1}{2} r \bar{\phi}^2
+    p(\bar{\phi}) \sim N e^{- \frac{1}{2} V r \bar{\phi}^2}
   \quad \text{for} \quad
@@
 
-In general,
-they depend on a renormalization scale $\mu$, and it is possible to derive an
-expression for $S_{\mathrm{eff}}$ evaluated at a generic value of $\mu$. Here,
-for simplicity we choose to evaluate (\ref{eq:effective_action}) at the scale
-$\mu^2 = r + p^2$. This value of $\mu$ is optimal for the reliability of
-perturbation theory, because it avoids the emergence of large logarithms.
+In order to describe how $p(\bar{\phi})$ depends on $R$ at fixed bare couplings
+$r_0$, $u_0$, it is necessary to account for renormalization effects. In
+general, the renormalized couplings $r$, $g$ depend on a renormalization scale
+$\mu$, and it is possible to derive an expression for $p(\bar{\phi})$ evaluated
+at a generic value of $\mu$. However, for simplicity,
+(\ref{eq:log_pdf}) is evaluated at the scale $\mu^2 = r + R^{-2}$.
+This choice of $\mu$ is optimal for the reliability of perturbation theory,
+because it avoids the emergence of large logarithms over the widest possible
+range of parameters.
 
-  As is well
-known, in $4 - \varepsilon$ dimensions it satisfies the renormalization group
-equation:
+The renormalization group equations for $r$ and $g$ are:
+\begin{equation}
+  \label{eq:renormalization_group}
+  \mu \frac{\dd g}{\dd \mu} = g^2\,,\quad
+  \mu \frac{\dd r}{\dd \mu} = \frac{1}{3}g r\,.
+\end{equation}
+
+These can be integrated and combined with the condition $\mu^2 = r + R^{-2}$ to
+obtain a system of equations\footnote{There are of course many alternative,
+arguably simpler, solutions that differ by subleading orders in an expansion in
+$g(R_1)$. The one displayed here is the exact solution to
+(\ref{eq:renormalization_group}).} connecting the renormalized couplings at two
+different values of $R$:
+\begin{equation}
+  \frac{g(R_1)}{g(R_2)} = \frac{r(R_1)^3}{r(R_2)^3} = 
+  1 - \frac{1}{2} g(R_1)\log\frac{r(R_2) + R_2^{-2}}{r(R_1) + R_1^{-2}}\,.
+\end{equation}
+
+\begin{figure}
+\begin{center}
+\includegraphics[scale=0.75]{renormalization_group.png}
+\end{center}
+    \caption{\label{fig:renormalization_group} A few solutions to
+    (\ref{eq:renormalization_group}) with $g(R_1) = 0.45$.  In the left plot,
+    the shaded region shows where the perturbative expansion becomes invalid.
+    Similarly in the right plot, the lines become dotted oustside of the
+    perturbative region. Note how the sign of $r$ is preserved, $r = 0$ is a
+    solution, and $g$ goes to $0$ for $R_2 \gg R_1$.}
+\end{figure}
+
+A few solutions to these equations are shown in
+fig.~\ref{fig:renormalization_group}. Within the perturbative regime $g \ll 1$,
+$r R^2 \gtrsim -1/2$, the coupling $r$ varies very little with $R$, and, as is
+clear from the differential form (\ref{eq:renormalization_group}), the sign of
+$r$ is always preserved. Thus we conclude that the critical point is at $r
+= 0$, and the symmetric phase is realized for $r > 0$.
+
+At the critical point, the renormalized coupling $g$ follows the simplified renormalization group equation:
+\begin{equation}
+    \frac{1}{g(R_2)} - \frac{1}{g(R_1)} = \log\frac{R_2}{R_1}\,,
+\end{equation}
+and hence, as system size grows, the renormalized coupling $g$ goes to zero.
+In this regime, the quartic term in (\ref{eq:log_pdf}) dominates all the
+others. This is perhaps most evident if the PDF is expressed in terms of the
+rescaled quantity $\bar{\varphi} = g^{\frac{1}{4}} \bar{\phi}$, whose
+variance remains finite as $g\to0$.
+
+Thus we conclude that, for sufficiently large system size: 
+\begin{equation}
+    \label{eq:log_pdf_critical_point}
+    -\log p(\phi) \sim \mathrm{const} + \frac{2\pi^2}{9} V g\bar{\phi}^4\,,
+\end{equation}
+with subleading terms vanishing like $1 / \log(R)$.
+
+In Monte Carlo simulations, the qualitative behavior of the distribution of the
+magnetization is often characterized by measuring the so-called Binder
+cumulant:
+\begin{equation}
+    U = 1 - \frac{\left\langle \bar \phi^4 \right \rangle}
+    {3\left\langle \bar \phi^2 \right \rangle^2},
+\end{equation}
+which is constructed to be independent of the overall scale of $\bar{\phi}$,
+and to be zero if $\bar{\phi}$ is normally distributed. From
+(\ref{eq:log_pdf_critical_point}) we conclude that, on a 4-Torus, at the
+critical point\footnote{ Incidentally, we note that this result is not in
+contraddiction with the claim in the original paper that the critical Binder
+cumulant in 4 dimensions is zero, because in the original paper the cumulant is
+not computed from the total magnetization, but rather from the average
+magnetization of blocks of spins much smaller than the overall system size.}:
+\begin{equation}
+    \label{eq:critical_binder_cumulant}
+    U = 1 - \frac{4\,\Gamma\left(\frac{5}{4}\right)^2}
+    {3\,\Gamma\left(\frac{3}{4}\right)^2} 
+    = 0.27052\ldots
+\end{equation}
+
+In fig.~\ref{fig:binder_cumulant}, we show a comparison of the Binder cumulant
+computed from (\ref{eq:log_pdf}) near the critical point, and the results of
+Monte Carlo simulation of the Ising model.
+
+\section{Derivation}
+
+We now describe briefly how the result (\ref{eq:log_pdf}) is obtained. First of
+all we observe that the average magnetization $\bar{\phi}$ is proportional to
+the zero momentum mode of the field $\phi$:
+\begin{equation}
+    \phi(x) = \frac{1}{V} \sum_{n \in \mathbb{Z}^4} \phi_{n} 
+    e^{i \frac{n\cdot x}{R}}\,;\quad \bar{\phi} = \frac{\phi_{0}}{V}\,.
+\end{equation}
+
+Thus, because of the delta function in (\ref{eq:pdf_definition}), the component
+$\phi_{0}$ becomes an external field, whereas the other components $\phi_{n}$
+with are still part of the functional integral. Separating the zero momentum
+mode from the other modes in the action yields:
+\begin{equation}
+\begin{split}
+    S(\bar{\phi}, \phi) =& 
+    V \left(\frac{1}{2} r_0 \bar{\phi}^2 + \frac{u_0}{24} \bar{\phi}^4\right) +
+    \frac{u_0}{4V} \bar{\phi}^2 \sum_{n} \phi_{n} \phi_{-n} + \\
+    &\frac{u_0}{6V^2} \bar{\phi} 
+        \sum_{n_1, n_2} \phi_{n_1} \phi_{n_2} \phi_{-n_1 -n_2} + \\
+    &\frac{u_0}{24V^3} \sum_{n_1, n_2, n_3} \phi_{n_1} \phi_{n_2} 
+        \phi_{n_3} \phi_{-n_1 -n_2 -n_3}\,.
+\end{split}
+\end{equation}
+
+We also introduce renormalized couplings
+\begin{equation}
+    u_0 = u(1 + u\delta u)\,;\quad r_0 = r(1 + u \delta r)\,,
+\end{equation}
+and we obtain the following edges and vertices in the diagrammatic expansion:
+\begin{displaymath}
+\begin{tikzpicture}[scale=0.5, baseline=(current bounding box.center)]
+    \draw[dashed, thick] (-1, -1) -- (0, 0) -- (-1, 1);
+    \draw[thick] (+1, -1) -- (0, 0) -- (+1, 1);
+    \node[above right] at (1, 1) {$n_1$};
+    \node[below right] at (1, -1) {$n_2$};
+\end{tikzpicture}
+    -\frac{u\bar{\phi}^2}{4V}\delta_{n_1 + n_2}
+    \quad\quad
+\begin{tikzpicture}[scale=0.5, baseline=(current bounding box.center)]
+    \draw[dashed, thick] (-1, -1) -- (0, 0);
+    \draw[thick] (0, 0) -- (-1, 1);
+    \draw[thick] (+1, -1) -- (0, 0) -- (+1, 1);
+    \node[above left] at (-1, 1) {$n_1$};
+    \node[above right] at (1, 1) {$n_2$};
+    \node[below right] at (1, -1) {$n_3$};
+\end{tikzpicture}
+    -\frac{u\bar{\phi}}{6V^2}\delta_{n_1 + n_2 + n_3}
+\end{displaymath}
+\begin{displaymath}
+\begin{tikzpicture}[scale=0.5, baseline=(current bounding box.center)]
+    \draw[thick] (-1, -1) -- (1, 1);
+    \draw[thick] (-1, 1) -- ( 1, -1);
+    \node[above left] at (-1, 1) {$n_1$};
+    \node[above right] at (1, 1) {$n_2$};
+    \node[below right] at (1, -1) {$n_3$};
+    \node[below left] at (-1, -1) {$n_4$};
+\end{tikzpicture}
+    -\frac{u}{24V^3}\delta_{n_1 + n_2 + n_3 + n_4}
+\quad\quad
+\begin{tikzpicture}[scale=0.5, baseline=(current bounding box.center)]
+    \draw[thick] (-1, 0) -- (1, 0);
+    \node[left] at (-1, 0) {$n_1$};
+    \node[right] at (1, 0) {$n_2$};
+\end{tikzpicture}
+    \quad
+    \frac{V}{n_1^2R^{-2} + r}\delta_{n_1 + n_2}
+\end{displaymath}
+plus additional vertices associated with the counterterms $\delta r$, $\delta
+u$ which we do not list for brevity.
+
+The logarithm of the probability distribution $p(\bar{\phi}$ is the sum of all connected diagrams:
+\begin{displaymath}
+\begin{split}
+    \log p(\bar{\phi}) = &\,
+    \mathrm{const} -V \left(\frac{1}{2} r \bar{\phi}^2 + 
+    \frac{u}{24} \bar{\phi}^4\right)\\
+&+\ 
+\begin{tikzpicture}[scale=0.5, baseline={([yshift=-.5ex]current bounding box.center)}]
+    \draw[thick, dashed] (-1, -0.5) -- (0, 0) -- (1, -0.5);
+    \draw[thick] (0, 0) .. controls (-1.5, 1.3) and (1.5, 1.3) .. (0, 0);
+\end{tikzpicture}
+\ + \ 
+\begin{tikzpicture}[scale=0.5, baseline={([yshift=-.5ex]current bounding box.center)}]
+    \draw[thick, dashed] (-1.75, 0) -- (-0.75, 0);
+    \draw[thick] (-0.75, 0) -- (0.75, 0);
+    \draw[thick, dashed] (1.75, 0) -- (0.75, 0);
+    \draw[thick] (0, 0) circle (0.75cm);
+\end{tikzpicture}
+\ + \ 
+\begin{tikzpicture}[scale=0.5, baseline={([yshift=-.5ex]current bounding box.center)}]
+    \draw[thick, dashed] (-1, -0.5) -- (0, 0) -- (1, -0.5);
+    \draw[thick] (0, 0) .. controls (0.6, 0.5).. (0, 1) .. 
+    controls (-1.5, 2.3) and (1.5, 2.3) .. (0, 1) .. 
+    controls (-0.6, 0.5) .. (0, 0) ;
+\end{tikzpicture}
+ \ + \ \cdots \\
+&+\ 
+\begin{tikzpicture}[scale=0.5, baseline={([yshift=-.5ex]current bounding box.center)}]
+    \draw[thick, dashed] (-1.75, -0.75) -- (-1, 0) -- (-1.75, 0.75);
+    \draw[thick, dashed] (1.75, -0.75) -- (1, 0) -- (1.75, 0.75);
+    \draw[thick] (-1, 0) .. controls (-0.2, 0.8) and (0.2, 0.8) .. (1, 0) .. 
+    controls (0.2, -0.8) and (-0.2, -0.8) .. (-1, 0);
+\end{tikzpicture}
+\ + \ 
+\begin{tikzpicture}[scale=0.5, baseline={([yshift=-.5ex]current bounding box.center)}]
+    \draw[thick, dashed] (-1.5, 0.75) -- (-0.75, 0) -- (-1.5, -0.75) ;
+    \draw[thick] (0, 0) circle (0.75cm);
+    \draw[thick, dashed] (0.53, 0.53) -- (1.25, 1.25);
+    \draw[thick, dashed] (0.53, -0.53) -- (1.25, -1.25);
+    \draw[thick] (0.53, -0.53) .. controls (0.2, -0.2) 
+    and (0.2, 0.2) .. (0.53, 0.53);
+\end{tikzpicture}
+\ + \ 
+\begin{tikzpicture}[scale=0.5, baseline={([yshift=-.5ex]current bounding box.center)}]
+    \draw[thick, dashed] (-1.75, -0.75) -- (-1, 0) -- (-1.75, 0.75);
+    \draw[thick, dashed] (1.75, -0.75) -- (1, 0) -- (1.75, 0.75);
+    \draw[thick] (-1, 0) .. controls (-0.2, 0.8) and (0.2, 0.8) .. (1, 0) .. 
+    controls (0.2, -0.8) and (-0.2, -0.8) .. (-1, 0);
+\end{tikzpicture}
+\end{split}
+\end{displaymath}
 
 \begin{equation}
-  \mu \frac{\dd g}{\dd \mu} = -\varepsilon g + g^2\,.
+\label{eq:f_functions}
+f_1 = 
 \end{equation}
 
-The Wilson-Fisher fixed point is attained at $g_{\star} = \varepsilon$, and
-substituting $g \to \varepsilon$ in (\ref{eq:effective_action}) and dropping
-higher order terms in $\varepsilon$ yields the effective action at the fixed
-point to leading order in $\epsilon$.
-
-The coupling $r$ satisfies the renormalization group equation:
-\begin{equation}
-  \mu \frac{\dd r}{\dd \mu} = \frac{g}{3}r
-\end{equation}
-
-$r$, $\eta$, and $g$ are renormalized couplings
-that satisfy the renormalization group equations:
-
-\begin{align}
-  &\mu \frac{\dd g}{\dd \mu} = -\varepsilon g + g^2 \\
-  &\mu \frac{\dd r}{\dd \mu} = \frac{g}{3}r \\
-  &\mu \frac{\dd \eta}{\dd \mu} = \frac{g}{3}(\eta - 2)
-\end{align}
+\end{document}
 
@@
 
-While a general expression for the effective action at a generic scale $\mu$ is
-available, for simplicity (\ref{eq:effective_action}) is evaluated at $\mu^2
-= r + p^2$. This value of $\mu$ is optimal for the reliability of perturbation
-theory, because it minimizes the effect of the so-called large logs.
+At the critical point it is convenient to introduce the dimensionless 
+quantity
+\begin{equation}
+    \bar{\varphi} =
+    \left(\frac{2\pi^2}{9} V g\right)^{\frac{1}{4}} \bar{\phi}\,,
+\end{equation}
+whose probability distribution is:
+\begin{equation}
+\label{eq:log_pdf_critical}
+\begin{split}
+    -\log p(\bar{\varphi}) = \mathrm{const.} +  
+    &g^{\frac{1}{2}}\bar{\varphi}^2 
+      \left(\frac{\pi}{\sqrt{2}}\left(f_1(0) - 1\right)  
+      + O\left(g\right) \right) + \\
+    &\bar{\varphi}^4\left(
+      1 - \frac{g}{2}f_2\left(0\right)
+      + O\left(g^2\right)\right) + \\
+    &g^{\frac{3}{2}} \bar{\varphi}^6 \left(\frac{f_3(0)}{3\sqrt{2} \pi}
+    + O\left(g\right)\right) + O\left(g^2\bar{\varphi}^8\right)\,.
+\end{split}
+\end{equation}
 
-An explicit expression for the functions $f_1$, $f_2$, $f_3$ is reported in the
-appendix. As shown in fig.~\ref{fig:f_functions}, $f_i(x)$ is of order 1 in for
-$x \gtrsim -1/2$, $f_1$ approaches a constant as $x \to \infty$, whereas
-$f_2$ and $f_3$ approach 0. They all diverge as $x$ approaches $-1$.
-\section{Appendix}
+This expression has a residual dependence on system size via dimensional
+transmutation of the renormalized coupling $g$, which at the critical point $r
+= 0$ follows the simplified renormalization group equation
+\begin{equation}
+    \frac{1}{g(R_2)} - \frac{1}{g(R_1)} = \log\frac{R_2}{R_1}\,.
+\end{equation}
 
+Therefore, as system size grows at fixed bare couplings $r_0$ $u_0$, the
+renormalized coupling $g$ goes to zero, and the all terms in
+(\ref{eq:log_pdf_critical}) besides the quartic vanish. Thus we conclude
+that, for sufficiently large system size: 
 \begin{equation}
-\label{eq:f_functions}
-f_1 = 
+    -\log p(\phi) \sim \mathrm{const} + \frac{2\pi^2}{9} V g\bar{\phi}^4\,.
 \end{equation}
-\end{document}

### 2020-03-08 ising_model/plot.py

@@ -109,7 +109,7 @@ def get_J_range(dim, mu):
     return J_range_table[(dim, mu)]
 
 
-def plot_time_series(file_name, max_length=8192):
+def plot_time_series_single(file_name, max_length, use_stamp,  axes):
     params = ising_model.udh.load_params(file_name)
     vol = 1
     for L in params.shape:
@@ -117,6 +117,7 @@ def plot_time_series(file_name, max_length=8192):
     observables_array = ising_model.udh.load_observables(file_name)
     time_series = collections.defaultdict(lambda: list())
     i = 0
+    stamps = []
     for observables in observables_array:
         #time_series['flip_cluster'].append(observables.flip_cluster_duration)
         #time_series['clear_flag'].append(observables.clear_flag_duration)
@@ -126,16 +127,25 @@ def plot_time_series(file_name, max_length=8192):
         time_series['phi2'].append(numpy.square(
             (observables.n_up - observables.n_down) / vol))
         time_series['n_holes'].append(observables.n_holes / vol)
+        stamps.append(observables.stamp)
         i += 1
         if i >= max_length:
             break
 
-    figure, axes = pyplot.subplots(len(time_series), 2, squeeze=False)
-    figure.set_size_inches(9, 4)
     for i, (key, value) in enumerate(time_series.items()):
-        axes[i, 0].plot(value, marker='o', linestyle='None', markersize=1)
+        if use_stamp:
+            abscissae = stamps
+        else:
+            abscissae = range(len(stamps))
+        axes[i, 0].plot(abscissae, value, marker='o', linestyle='None', markersize=1)
         axes[i, 0].set_title(key)
-        axes[i, 1].hist(value, bins=round(len(value) / 64))
+        axes[i, 1].hist(value, bins=32, alpha=0.5)
+
+
+def plot_time_series(file_names, max_length=8192, use_stamp=False):
+    figure, axes = pyplot.subplots(2, 2, squeeze=False)
+    figure.set_size_inches(9, 4)
+    for file_name in file_names:
+        plot_time_series_single(file_name, max_length, use_stamp, axes)
     figure.tight_layout()
     return figure, axes
-

### 2020-03-08 ising_model/scripts/update-file-groups.py

@@ -53,7 +53,7 @@ def print_table(name, fields, stream):
         print_data(fields["data"], "{:4}, {:4}", stream)
     elif name == "reweight_args":
         print_data(fields["data"],
-                   "{:4}, {:9.6f}, {:9.6f}, {:9.6f}, {:4}", stream)
+                "{:4}, {:9.6f}, {:9.7f}, {:9.7f}, {:4}, {:4}", stream)
     stream.write("    ]\n")
     stream.write("  }")
 

### 2020-03-20 paper/paper.tex
Paper checkpoint

@@
     &\frac{u_0}{24V^3} \sum_{n_1, n_2, n_3} \phi_{n_1} \phi_{n_2} 
-        \phi_{n_3} \phi_{-n_1 -n_2 -n_3}\,.
+        \phi_{n_3} \phi_{-n_1 -n_2 -n_3}\,,
 \end{split}
 \end{equation}
+where all summations now exclude $n = (0, 0, 0, 0)$.
 
@@
 \begin{tikzpicture}[scale=0.5, baseline=(current bounding box.center)]
-    \draw[dashed, thick] (-1, -1) -- (0, 0) -- (-1, 1);
+    \draw[densely dashed, thick] (-1, -1) -- (0, 0) -- (-1, 1);
     \draw[thick] (+1, -1) -- (0, 0) -- (+1, 1);
@@
 \begin{tikzpicture}[scale=0.5, baseline=(current bounding box.center)]
-    \draw[dashed, thick] (-1, -1) -- (0, 0);
+    \draw[densely dashed, thick] (-1, -1) -- (0, 0);
     \draw[thick] (0, 0) -- (-1, 1);
@@
 
-The logarithm of the probability distribution $p(\bar{\phi}$ is the sum of all connected diagrams:
+The logarithm of the probability distribution of $\bar{\phi}$ is the sum of all
+connected diagrams:
 \begin{displaymath}
@@
 \begin{tikzpicture}[scale=0.5, baseline={([yshift=-.5ex]current bounding box.center)}]
-    \draw[thick, dashed] (-1, -0.5) -- (0, 0) -- (1, -0.5);
-    \draw[thick] (0, 0) .. controls (-1.5, 1.3) and (1.5, 1.3) .. (0, 0);
+    \draw[thick, densely dashed] (-1.75, -0.75) -- (-1, 0) -- (-1.75, 0.75);
+    \draw[thick] (-1, 0) .. controls (-0.2, 0.8) and (0.5, 0.8) .. (0.5, 0) .. 
+    controls (0.5, -0.8) and (-0.2, -0.8) .. (-1, 0);
 \end{tikzpicture}
@@
 \begin{tikzpicture}[scale=0.5, baseline={([yshift=-.5ex]current bounding box.center)}]
-    \draw[thick, dashed] (-1.75, 0) -- (-0.75, 0);
+    \draw[thick, densely dashed] (-1.75, 0) -- (-0.75, 0);
     \draw[thick] (-0.75, 0) -- (0.75, 0);
-    \draw[thick, dashed] (1.75, 0) -- (0.75, 0);
+    \draw[thick, densely dashed] (1.75, 0) -- (0.75, 0);
     \draw[thick] (0, 0) circle (0.75cm);
@@
 \begin{tikzpicture}[scale=0.5, baseline={([yshift=-.5ex]current bounding box.center)}]
-    \draw[thick, dashed] (-1, -0.5) -- (0, 0) -- (1, -0.5);
-    \draw[thick] (0, 0) .. controls (0.6, 0.5).. (0, 1) .. 
-    controls (-1.5, 2.3) and (1.5, 2.3) .. (0, 1) .. 
-    controls (-0.6, 0.5) .. (0, 0) ;
+    \draw[thick, densely dashed] (-2.75, -0.75) -- (-2, 0) -- (-2.75, 0.75);
+    \draw[thick] (-2, 0) .. controls (-1.2, 0.8) and (-0.8, 0.8) .. (0, 0) .. 
+    controls (-0.8, -0.8) and (-1.2, -0.8) .. (-2, 0);
+    \draw[thick] (1.6, 0) .. controls (1.6, 0.8) and (0.8, 0.8) .. (0, 0) .. 
+    controls (0.8, -0.8) and (1.6, -0.8) .. (1.6, 0);
 \end{tikzpicture}
@@
 \begin{tikzpicture}[scale=0.5, baseline={([yshift=-.5ex]current bounding box.center)}]
-    \draw[thick, dashed] (-1.75, -0.75) -- (-1, 0) -- (-1.75, 0.75);
-    \draw[thick, dashed] (1.75, -0.75) -- (1, 0) -- (1.75, 0.75);
+    \draw[thick, densely dashed] (-1.75, -0.75) -- (-1, 0) -- (-1.75, 0.75);
+    \draw[thick, densely dashed] (1.75, -0.75) -- (1, 0) -- (1.75, 0.75);
     \draw[thick] (-1, 0) .. controls (-0.2, 0.8) and (0.2, 0.8) .. (1, 0) .. 
@@
 \begin{tikzpicture}[scale=0.5, baseline={([yshift=-.5ex]current bounding box.center)}]
-    \draw[thick, dashed] (-1.5, 0.75) -- (-0.75, 0) -- (-1.5, -0.75) ;
+    \draw[thick, densely dashed] (-1.5, 0.75) -- (-0.75, 0) -- (-1.5, -0.75) ;
     \draw[thick] (0, 0) circle (0.75cm);
-    \draw[thick, dashed] (0.53, 0.53) -- (1.25, 1.25);
-    \draw[thick, dashed] (0.53, -0.53) -- (1.25, -1.25);
+    \draw[thick, densely dashed] (0.53, 0.53) -- (1.25, 1.25);
+    \draw[thick, densely dashed] (0.53, -0.53) -- (1.25, -1.25);
     \draw[thick] (0.53, -0.53) .. controls (0.2, -0.2) 
@@
 \begin{tikzpicture}[scale=0.5, baseline={([yshift=-.5ex]current bounding box.center)}]
-    \draw[thick, dashed] (-1.75, -0.75) -- (-1, 0) -- (-1.75, 0.75);
-    \draw[thick, dashed] (1.75, -0.75) -- (1, 0) -- (1.75, 0.75);
-    \draw[thick] (-1, 0) .. controls (-0.2, 0.8) and (0.2, 0.8) .. (1, 0) .. 
-    controls (0.2, -0.8) and (-0.2, -0.8) .. (-1, 0);
+    \draw[thick, densely dashed] (-2.75, -0.75) -- (-2, 0) -- (-2.75, 0.75);
+    \draw[thick, densely dashed] (2.75, -0.75) -- (2, 0) -- (2.75, 0.75);
+    \draw[thick] (-2, 0) .. controls (-1.2, 0.8) and (-0.8, 0.8) .. (0, 0) .. 
+    controls (-0.8, -0.8) and (-1.2, -0.8) .. (-2, 0);
+    \draw[thick] (2, 0) .. controls (1.2, 0.8) and (0.8, 0.8) .. (0, 0) .. 
+    controls (0.8, -0.8) and (1.2, -0.8) .. (2, 0);
 \end{tikzpicture}
+ \ + \ \cdots \\
+& + \ 
+\begin{tikzpicture}[scale=0.5, baseline={([yshift=-.5ex]current bounding box.center)}]
+    \draw[thick, densely dashed] (-1.5, 0.75) -- (-0.75, 0) -- (-1.5, -0.75) ;
+    \draw[thick] (0, 0) circle (0.75cm);
+    \draw[thick, densely dashed] (1.40, -0.92) -- (0.375, -0.65) -- (0.1, -1.67);
+    \draw[thick, densely dashed] (1.40, 0.92) -- (0.375, 0.65) -- (0.1, 1.67);
+\end{tikzpicture}
+ \ + \ \cdots \,,
 \end{split}
 \end{displaymath}
+where again we left out all diagrams involving the counterterms for brevity.
+
+Retaining only one-loop diagrams, and employing a heath kernel regulator we have:
+\begin{equation}
+\begin{split}
+    -\log p(\bar{\phi}) = \mathrm{const} + V \Bigg( 
+    &\frac{1}{2} \bar{\phi}^2\left( r + u \left(\delta r + \frac{1}{2} I_1\right) + O(u^2)\right) + \\
+    &\frac{u}{24} \bar{\phi}^4 \left(1 + u \left(\delta u - \frac{3}{2} I_2\right) + O(u^2)\right) + \\
+    &\frac{u^3}{48}\bar{\phi}^6\left(I_3 + O(u)\right) + O(u^4 \bar{\phi}^8) \Bigg)\,,
+\end{split}
+\end{equation}
+where
+\begin{equation}
+    I_k = \frac{1}{V} \sum_{n\in\mathbb{Z}^4\setminus\{0\}} 
+    \frac{e^{-s(r + n^2 R^{-2})}}{\left(r + n^2 R^{-2}\right)^k} = 
+    \frac{1}{(k - 1)!}\left(-\frac{\partial}{\partial r} - s\right)^k I_1
+\end{equation}
+
+The summation $I_1$ can be simplified substantially using the following trick:
+\begin{equation}
+\begin{split}
+    I_1 &= \frac{1}{V} \int_{s}^{\infty} \dd t \sum_{n\in\mathbb{Z}^4\setminus\{0\}} 
+    e^{-t(r + n^2 R^{-2})}\\
+    & = \frac{1}{V} \int_{s}^{\infty} \dd t\ e^{-t r}
+        \left(\left(\sum_{m = -\infty}^{\infty} e^{-t m^2 R^{-2}}\right)^4 - 1\right) \\
+    & = \frac{1}{V} \int_{s}^{\infty} \dd t\ e^{-t r}
+        \left(\theta\!\left(t R^{-2}\right)^4 - 1\right)\,,
+\end{split}
+\end{equation}
+where:\footnote{The function $\theta$ is in fact a Jacobi theta function, but we
+won't need any of its special properties beyond its asymptotic behavior, which
+can be easily obtained from the definition.}
+\begin{equation}
+    \theta(z) \equiv \sum_{m = -\infty}^{\infty} e^{-m^2 z}\,.
+\end{equation}
+
+We now extract the divergent part of $I_1$ by subtracting under the integral a
+function that has the same asymptotic behavior as the integrand for $t \to 0$,
+but whose integral can be computed in closed form. In order to do that, we need
+the asymptotic behavior of $\theta(z)$ for $z\to0$, which can be obtained from
+its definition using the Euler-Mclaurin formula:
+\begin{equation}
+    \theta(z) \sim \sqrt{\frac{\pi}{z}} + o(z^k) \quad\mathrm{for}\, z\to0\,.
+\end{equation}
+
+Thus we have:
+\begin{equation}
+\begin{split}
+    I_1 &= \frac{1}{16\pi^2}\left(\int_{s}^{\infty}\dd t\, e^{-t \mu^2} t^{-2}\left(1 + t (\mu^2 - r)\right)
+    +\frac{1}{R^2} f(r R^2, \mu R)\right)\\
+    &= \frac{1}{16\pi^2}\left(\frac{1}{s} + r\left(\log s\mu^2 + \gamma_{E}\right) - 
+    \mu^2 +\frac{1}{R^2} f(r R^2, \mu R)\right)\,,
+\end{split}
+\end{equation}
+where $\mu$ is a renormalization scale that can be chosen at will, and:
+\begin{equation}
+    f(rR^2,\mu R) = \int_{0}^{\infty} \dd z\ e^{-z r R^2} \left(\frac{\theta(z)^4 - 1}{\pi^2} - 
+    \frac{1 + z\left(\mu^2 - r\right)R^2}{t^2}e^{-z\left(\mu^2 - r\right)R^2}\right).
+\end{equation}
 

### 2020-03-22 paper/paper.tex
Paper checkpoint

@@
 \end{center}
-\caption{\label{fig:binder_cumulant} Blah Blah}
+    \caption{\label{fig:binder_cumulant} Binder cumulant of the total
+    magnetization of the 4D Ising model. The left plot gives a broad picture,
+    the right plot shows the critical region close-up.  The shaded regions
+    display the one-sigma confidence intervals obtained from Monte Carlo
+    simulation. The center lines are obtained from a 3-parameter fit of the
+    perturbative result (\ref{eq:log_pdf}). The horizontal dashed line
+    indicates the critical value (\ref{eq:critical_binder_cumulant}). The fit
+    has $J_c = 0.1496938$, $J - J_c = -0.027\cdot r$, $g = 0.43$, with $r$, $g$
+    defined at the renormalization scale $L = 2\pi R = 8$. }
 \end{figure}
@@
 magnetization is often characterized by measuring the so-called Binder
-cumulant:
+cumulant \cite{Binder1981}:
 \begin{equation}
@@
 critical point\footnote{ Incidentally, we note that this result is not in
-contraddiction with the claim in the original paper that the critical Binder
-cumulant in 4 dimensions is zero, because in the original paper the cumulant is
-not computed from the total magnetization, but rather from the average
-magnetization of blocks of spins much smaller than the overall system size.}:
+contraddiction with the claim in the original paper \cite{Binder1981} that the
+critical Binder cumulant in 4 dimensions is zero, because in the original paper
+the cumulant is not computed from the total magnetization, but rather from the
+average magnetization of blocks of spins much smaller than the overall system
+size.}:
 \begin{equation}
@@
 computed from (\ref{eq:log_pdf}) near the critical point, and the results of
-Monte Carlo simulation of the Ising model.
+Monte Carlo simulation of the $4D$ Ising model. The agreement is excellent except
+for the smallest system size $L = 8$, probably because of contributions from
+irrelevant operators. Notice how slowly the finite size Binder cumulant
+approaches the asymptotic value (\ref{eq:critical_binder_cumulant}).
 
@@
 
-We now describe briefly how the result (\ref{eq:log_pdf}) is obtained. First of
-all we observe that the average magnetization $\bar{\phi}$ is proportional to
-the zero momentum mode of the field $\phi$:
+We now describe briefly how the result (\ref{eq:log_pdf}) is obtained. The
+perturbative approach is very similar to the computation of the effective
+action, as in \textit{e.g.} \cite{WeinbergEffectiveAction}, except that we are
+interested in the whole probability distribution of $\bar{\phi}$, instead of
+just the expected value. The main difficulty lies in evaluating the loop
+integrals at finite size.
+
+First of all we observe that the average magnetization
+$\bar{\phi}$ is proportional to the zero momentum mode of the field $\phi$:
 \begin{equation}
@@
 \begin{equation}
+    \label{eq:theta}
     \theta(z) \equiv \sum_{m = -\infty}^{\infty} e^{-m^2 z}\,.
@@
 
-We now extract the divergent part of $I_1$ by subtracting under the integral a
-function that has the same asymptotic behavior as the integrand for $t \to 0$,
-but whose integral can be computed in closed form. In order to do that, we need
-the asymptotic behavior of $\theta(z)$ for $z\to0$, which can be obtained from
-its definition using the Euler-Mclaurin formula:
+We now extract the divergent and finite parts of $I_1$ as $s\to 0$ by
+subtracting under the integral a function that has the same asymptotic behavior
+as the integrand for $t \to 0$, but whose integral can be computed in closed
+form. In order to do that, we need the asymptotic behavior of $\theta(z)$ for
+$z\to0$, which can be obtained from its definition using the Euler-Mclaurin
+formula:
 \begin{equation}
@@
     I_1 &= \frac{1}{16\pi^2}\left(\int_{s}^{\infty}\dd t\, e^{-t \mu^2} t^{-2}\left(1 + t (\mu^2 - r)\right)
-    +\frac{1}{R^2} f(r R^2, \mu R)\right)\\
+    +\frac{1}{R^2} f(r R^2, \mu R) + O(s)\right)\\
     &= \frac{1}{16\pi^2}\left(\frac{1}{s} + r\left(\log s\mu^2 + \gamma_{E}\right) - 
-    \mu^2 +\frac{1}{R^2} f(r R^2, \mu R)\right)\,,
+    \mu^2 +\frac{1}{R^2} f(r R^2, \mu R) + O(s)\right)\,,
 \end{split}
@@
 \begin{equation}
-    f(rR^2,\mu R) = \int_{0}^{\infty} \dd z\ e^{-z r R^2} \left(\frac{\theta(z)^4 - 1}{\pi^2} - 
-    \frac{1 + z\left(\mu^2 - r\right)R^2}{t^2}e^{-z\left(\mu^2 - r\right)R^2}\right).
+    \label{eq:f_definition}
+    f(\bar{r},\bar{\mu}) = \int_{0}^{\infty} \dd z\ e^{-z \bar{r}} \left(\frac{\theta(z)^4 - 1}{\pi^2} - 
+    \frac{1 + z\left(\bar{\mu}^2 - \bar{r}\right)}{z^2}e^{-z\left(\bar{\mu}^2 - \bar{r}\right)}\right).
 \end{equation}
 
+We set the counterterms to:
+\begin{align}
+    &\delta r = -\frac{1}{32\pi^2}\left(\frac{1}{s} + r\left(\log s \mu^2 + \gamma_E - 1\right)\right)\,,\\
+    &\delta u = -\frac{3}{32\pi^2}\left(\log s\mu^2 +\gamma_E + 1\right)\,,
+\end{align}
+from which the renormalization group equations (\ref{eq:renormalization_group})
+can be obtained, and we have:
 \begin{equation}
+\begin{split}
+    -\log p(\bar{\phi}) =& \mathrm{const}+ V \Bigg( \\
+    &\frac{1}{2} \bar{\phi}^2\left( r + \frac{u}{32\pi^2}  \left(\left(\frac{1}{R^2} f(rR^2, \mu R) + r - \mu^2\right)\right) + O(u^2)\right) + \\
+    &\frac{u}{24} \bar{\phi}^4 \left(1 + \frac{3u}{32\pi^2} f^{(1, 0)}(rR^2, \mu R) + O(u^2)\right) + \\
+    &\frac{u^3}{48}\bar{\phi}^6\left(\frac{R^2}{32\pi^2}f^{(2, 0)}(rR^2, \mu R) + O(u)\right) + O(u^4 \bar{\phi}^8) \Bigg)\,,
+\end{split}
+\end{equation}
+from which (\ref{eq:log_pdf}) is obtained by setting $g = \frac{3 u}{16
+\pi^2}$ and $\mu^2 = r + R^{-2}$. This last choice is motivated as follows. The
+subtraction in (\ref{eq:f_definition}) is similar to:
+\begin{equation}
+    \int_0^{\infty} \dd z\, \frac{e^{-a z} - e^{-b z}}{z} = \log \frac{b}{a}\,;
+\end{equation}
+when the asymptotic behavior of the two terms for large $z$ is not well
+matched, the integral becomes large in magnitude, making the perturbative
+expansion less reliable. With the choice $\mu^2 = r + R^{-2}$, both terms in
+(\ref{eq:f_definition}) have the same asymptotic behavior $\sim e^{-z(\bar{r} +
+1)}$, thus avoiding the large log problem over the widest possible range of
+parameters.
+
+For completeness, let us display explicitly the functions $f_i$ that
+parametrize (\ref{eq:log_pdf}):
+\begin{align}
 \label{eq:f_functions}
-f_1 = 
-\end{equation}
+    &f_1(\bar r) = f(\bar r, \sqrt{\bar r + 1}) = 
+    \int_{0}^{\infty} \dd z\ e^{-z \bar{r}} \left(\frac{\theta(z)^4 - 1}{\pi^2} - 
+    \frac{(1 + z) e^{-z}}{z^2}\right)\,,\\
+    &f_2(\bar r) = f^{(1, 0)}(\bar r, \sqrt{\bar r + 1}) = 
+    \int_{0}^{\infty} \dd z\ e^{-z \bar{r}} z \left(\frac{\theta(z)^4 - 1}{\pi^2} - 
+    \frac{e^{-z}}{z^2}\right)\,,\\
+    &f_3(\bar r) = f^{(2, 0)}(\bar r, \sqrt{\bar r + 1}) = 
+    \int_{0}^{\infty} \dd z\ e^{-z \bar{r}} z^2 \frac{\theta(z)^4 - 1}{\pi^2}\,,
+\end{align}
+and $\theta$ is given by (\ref{eq:theta}).
+
+\begin{thebibliography}{9}
+\bibitem{Binder1981}
+K. Binder,
+\textit{Critical Properties from Monte Carlo Coarse Graining and Renormalization},
+\href{https://doi.org/10.1103/PhysRevLett.47.693}{Phys. Rev. Lett. \textbf{47} (1981) 693}.
+
+\bibitem{WeinbergEffectiveAction}
+S. Weinberg,
+\textit{The quantum theory of fields, Vol. 2}, Cambridge University Press, (1996) pg. 63.
+\end{thebibliography}
 

### 2020-03-22 paper/paper.tex
Paper checkpoint

@@
 \title{One-loop PDF of the average Ising magnetization in 4 dimensions.}
-\author{Andrea Allais}
+\author[a]{Andrea Allais}
+\affiliation[a]{Independent scholar}
 \emailAdd{allais.andrea@gmail.com}
-
+\abstract{We compute at one loop in perturbation theory the probability density
+function of the average magnetization $\bar{\phi}$ of the Ising model on both
+the 4-torus and the 4-sphere. The perturbative expansion covers the entire
+symmetric phase and a neighborhood of the critical point. We find that, at the
+critical point, the PDF has the asymptotic behavior $p(\bar{\phi})\sim
+\exp(-f(R) \bar{\phi}^4)$, and hence the critical value of the Binder cumulant
+is $U = 1 - \frac{4\,\Gamma(5/4)^2}{ 3\,\Gamma(3/4)^2}$. We validate our
+results by comparison with Monte Carlo simulation.}
 \begin{document}
@@
 \end{center}
-\caption{\label{fig:f_functions} The functions $f_i$}
+\caption{\label{fig:f_functions} The functions $f_i$ that appear in
+    (\ref{eq:log_pdf}). The functions go to zero as $x\to\infty$, and diverge
+    for $x\to-1$ ($x\to-4$ for the sphere), signaling an instability of the
+    perturbative vacuum.}
 \end{figure}
@@
 
-\section{Derivation}
+\section{Derivation on the 4-torus}
 
@@
 \end{equation}
-where:\footnote{The function $\theta$ is in fact a Jacobi theta function, but we
-won't need any of its special properties beyond its asymptotic behavior, which
-can be easily obtained from the definition.}
+where:
 \begin{equation}
     \label{eq:theta}
-    \theta(z) \equiv \sum_{m = -\infty}^{\infty} e^{-m^2 z}\,.
+    \theta(z) = \theta_3(0, e^{-z}) = \sum_{m = -\infty}^{\infty} e^{-m^2 z}\,.
 \end{equation}
+Here $\theta_3$ is a Jacobi theta function, but we will not need any of
+its special properties beyond its asymptotic behavior, which can be easily
+obtained from the definition.
 
@@
 
+\section{Results for the 4-sphere}
+\label{sec:four_sphere}
+It is possible to obtain $\log p(\bar{\phi})$ on the 4 sphere as well, with
+very similar methods. Here we highlight the main differences from the 4 torus.
+
+Curved manifolds allow for an additional renormalizable coupling: 
+\begin{equation}
+    S += \int \dd^4 x\ \frac{1}{2}\xi_0 \phi^2 \mathcal{R}\,,
+\end{equation}
+where $\mathcal{R}$ is the scalar curvature and $\xi_0$ is a dimensionless bare
+coupling. On a static manifold, this coupling is equivalent to a mass term.
+The free theory is Weyl invariant if $\xi_0 = \frac{1}{6}$. In the
+presence of interactions, the coupling needs to be renormalized as well, and
+its renormalized counterpart $\xi$ becomes a running coupling. At one loop, the
+renormalization group equation for $\xi$ is:
+\begin{equation}
+    \mu \frac{\dd \xi}{\dd \mu} =  \frac{1}{3} g \left(\xi - \frac{1}{6}\right)\,.
+\end{equation}
+
+From this experssion it seems that $\xi$ can be set to the critical value
+$\frac{1}{6}$ at all energy scales. However, this turns out to be an illusion:
+at higher orders in perturbation theory the renormalization group equation
+becomes inhomogeneous \cite{Brown1980}. Therefore, on a curved manifold, $\xi$
+is simply another free parameter of the scalar field theory.
+
+The probability distribution of the total magnetization on the 4-sphere is
+still given by (\ref{eq:log_pdf}), with the substitution:
+\begin{equation}
+    rR^2 \to \frac{\xi}{12} + r R^2\,
+\end{equation}
+where $R$ is now the radius of the sphere, and with the functions $f_i$ defined
+as:
+\begin{align}
+    &f_1(\bar{r}) = (2 - \bar{r})\int \dd z\ e^{-z(\bar{r} + 4)} \left(H(z) - \frac{1}{z}\right) + \frac{7}{3}\,,\\
+    &f_2(\bar{r}) = \int \dd z\ e^{-z(\bar{r} + 4)} 
+    \left(1 + z (2 - \bar{r})\right) \left(H(z) - \frac{1}{z}\right)+ \frac{6}{\bar{r} + 4}\,,\\
+    &f_3(\bar{r}) = \int \dd z\ e^{-z(\bar{r} + 4)} 
+    \left(2 + z (2 - \bar{r})\right)z \left(H(z) - \frac{1}{z}\right) + \frac{\bar{r} + 10 }{(\bar{r} + 4)^2}\,,\\
+\end{align}
+where:
+\begin{equation}
+    H(z) = \sum_{\ell = 1}^{\infty} (2\ell + 3)e^{-z (\ell(\ell + 3) - 4)}\,.
+\end{equation}
+
+The functions $f_i$ for the sphere are also displayed in
+fig.~\ref{fig:f_functions}. They diverge for $\bar{r} \to -4$, signaling the
+instability of the perturbative vacuum, and they go to zero for $\bar{r} \to
+\infty$.
+
 \begin{thebibliography}{9}
@@
 \textit{The quantum theory of fields, Vol. 2}, Cambridge University Press, (1996) pg. 63.
+
+\bibitem{Brown1980}
+Lowell S. Brown, John C. Collins,
+\textit{Dimensional renormalization of scalar field theory in curved space-time},
+\href{https://doi.org/10.1016/0003-4916(80)90232-8}{Ann. Phys. \textbf{130} (1980) 215}.
 \end{thebibliography}
@@
 where the renormalization group equation becomes inhomogeneous.
-
-At the critical point it is convenient to introduce the dimensionless 
-quantity
-\begin{equation}
-    \bar{\varphi} =
-    \left(\frac{2\pi^2}{9} V g\right)^{\frac{1}{4}} \bar{\phi}\,,
-\end{equation}
-whose probability distribution is:
-\begin{equation}
-\label{eq:log_pdf_critical}
-\begin{split}
-    -\log p(\bar{\varphi}) = \mathrm{const.} +  
-    &g^{\frac{1}{2}}\bar{\varphi}^2 
-      \left(\frac{\pi}{\sqrt{2}}\left(f_1(0) - 1\right)  
-      + O\left(g\right) \right) + \\
-    &\bar{\varphi}^4\left(
-      1 - \frac{g}{2}f_2\left(0\right)
-      + O\left(g^2\right)\right) + \\
-    &g^{\frac{3}{2}} \bar{\varphi}^6 \left(\frac{f_3(0)}{3\sqrt{2} \pi}
-    + O\left(g\right)\right) + O\left(g^2\bar{\varphi}^8\right)\,.
-\end{split}
-\end{equation}
-
-This expression has a residual dependence on system size via dimensional
-transmutation of the renormalized coupling $g$, which at the critical point $r
-= 0$ follows the simplified renormalization group equation
-\begin{equation}
-    \frac{1}{g(R_2)} - \frac{1}{g(R_1)} = \log\frac{R_2}{R_1}\,.
-\end{equation}
-
-Therefore, as system size grows at fixed bare couplings $r_0$ $u_0$, the
-renormalized coupling $g$ goes to zero, and the all terms in
-(\ref{eq:log_pdf_critical}) besides the quartic vanish. Thus we conclude
-that, for sufficiently large system size: 
-\begin{equation}
-    -\log p(\phi) \sim \mathrm{const} + \frac{2\pi^2}{9} V g\bar{\phi}^4\,.
-\end{equation}

### 2020-04-18 paper/paper.tex
Proofread 1

@@
 
-\title{One-loop PDF of the average Ising magnetization in 4 dimensions.}
+\title{Perturbative probability distribution of the average Ising magnetization
+in 4 dimensions.}
+
 \author[a]{Andrea Allais}
@@
 \emailAdd{allais.andrea@gmail.com}
-\abstract{We compute at one loop in perturbation theory the probability density
-function of the average magnetization $\bar{\phi}$ of the Ising model on both
+
+\abstract{We compute, at one loop in perturbation theory, the probability density
+function of the average magnetization $\bar{\phi}$ of the Ising model, on both
 the 4-torus and the 4-sphere. The perturbative expansion covers the entire
@@
 critical point, the PDF has the asymptotic behavior $p(\bar{\phi})\sim
-\exp(-f(R) \bar{\phi}^4)$, and hence the critical value of the Binder cumulant
+\exp(-c \bar{\phi}^4)$, and hence the critical value of the Binder cumulant
 is $U = 1 - \frac{4\,\Gamma(5/4)^2}{ 3\,\Gamma(3/4)^2}$. We validate our
@@
 The average magnetization $\bar{\phi} = \frac{1}{V} \sum_{i} \sigma_i$ of the
-Ising model is the simplest observable in Monte Carlo simulations. For large
-system size, the probability distribution of $\bar{\phi}$ changes qualitatively
-between the two phases of the model. In the symmetric phase, if the correlation
-length $\xi$ is large in lattice units, but small compared to the system size
-$L$, the distribution is well approximated by a normal distribution centered at
-0. In the broken symmetry phase, still for $1 \ll \xi \ll L$, the distribution
-is well approximated by a bimodal superposition of two normals centered at
+Ising model is the simplest observable in Monte Carlo simulations. When the
+system size $L$ is large, the probability distribution of $\bar{\phi}$ changes
+qualitatively between the two phases of the model. In the symmetric phase,
+\textit{i. e.} when the correlation length $\xi$ is large in lattice units, but
+small compared to the system size, the distribution is well approximated by a
+zero-mean normal. In the broken symmetry phase, still for $1 \ll \xi \ll L$,
+the distribution is well approximated by a mixture of two normals centered at
 non-zero values $\pm \bar{\phi}_0$.
 
-Near the critical point $1 \ll L \ll \xi$ the probability distribution of
-$\bar{\phi}$ is a non-trivial function of $\bar{\phi}$, at least in two and
-three dimensions.  It is not immediately clear if the same is true in four
-dimensions as well.  Since the $\phi^4$ field theory that describes the Ising
-critical point in four dimensions becomes weakly coupled in the infrared, one
-may expect the probability distribution to be a single zero-centered normal at
-the critical point as well.
+Near the critical point, \textit{ie.} $1 \ll L \ll \xi$, the probability
+distribution of $\bar{\phi}$ is a non-trivial function of $\bar{\phi}$, at
+least in two and three dimensions.  It is not immediately clear if the same is
+also true in four dimensions. Since the field theory that describes the Ising
+critical point in four dimensions becomes weakly coupled at low energy, one may
+expect the probability distribution of $\bar{\phi}$ to remain a zero-centered
+normal at the critical point as well as in the symmetric phase.
 
-To clarify this issue, we compute the log pdf of $\bar{\phi}$ at one-loop in
-$4$ dimensions, and show that for very large system size, the critical
-probability distribution is not normal, but rather of the form
-$-\log p(\bar{\phi})\sim f(L) \bar{\phi}^4$. We find that corrections to this
-form vanish very slowly with increasing system size, like $1/\log L$.
+To clarify this issue, we compute the logarithm of the PDF of $\bar{\phi}$ at
+one-loop in $4$ dimensions, and show that, for very large system size, the
+critical probability distribution is not normal, but rather of the form $-\log
+p(\bar{\phi})\sim f(L) \bar{\phi}^4$. We find that corrections to this form
+vanish very slowly with increasing system size, like $1/\log L$.
 
@@
   S[\phi] \equiv \int \dd^4 x\ \left(
-    \frac{1}{2} \phi\left(\Delta + r_0\right)\phi +
-    \frac{1}{24} u_0 \phi^4\right)
+    \frac{1}{2} \phi(x)\left(\Delta + r_0\right)\phi(x) +
+    \frac{1}{24} u_0 \phi(x)^4\right)
 \end{equation}
 defined on a 4-torus of radius $R$, i.e. $x_i\sim x_i + 2\pi R$. In this field
-theory context, the pdf of the average magnetization is defined as:
+theory context, the PDF of the average magnetization is defined as:
 \begin{equation}
@@
     p(\bar\phi) \equiv \frac{1}{Z} \int \left[\mathcal{D}\phi\right]
-    \delta\left(\bar{\phi} - \frac{1}{V}\int\dd^4 x\ \phi\right)
+    \delta\left(\bar{\phi} - \frac{1}{V}\int\dd^4 x\ \phi(x)\right)
     e^{-S[\phi]}\,,
@@
 and $u_0$. An explicit expression for the functions $f_1$, $f_2$ and $f_3$ is
-reported in the appendix (\ref{eq:f_functions}), and they are plotted in
+reported in section \ref{sect:derivation} (\ref{eq:f_functions}), and they are
+plotted in
 fig.~\ref{fig:f_functions}.
@@
 Eq. (\ref{eq:log_pdf}) is our central result, and we now describe it
-at lenght. We first illustrate its domain of validity and its dependence on the
+at length. We first illustrate its domain of validity and its dependence on the
 renormalized couplings $r$, $g$ on a torus of fixed radius $R$, then describe
@@
 The dimension-2 coupling $r$ controls the cross-over between the symmetric and
-broken-symmetry phases. The expression (\ref{eq:log_pdf}) is valid for
-all positive values of $r$, and also for negative values, provided that $rR^2
-\gtrsim -1/2$.  For lower values of $r$, additional modes of the laplacian
+broken-symmetry phases. The expression (\ref{eq:log_pdf}) is valid for all
+positive values of $r$, and also for negative values, provided that $rR^2
+\gtrsim -1/2$.  For lower values of $r$, additional modes of the Laplacian
 condense beside the zero-mode, and this invalidates the perturbative expansion.
-. This becomes manifest in a divergence of the functions $f_i$ as their
-argument approaches -1.  On the other hand, $f_1$, $f_2$, $f_3$ all go to zero
-as their argument approaches positive infinity, and hence for large $r$ the
-probability distribution of $\bar{\phi}$ is a zero-centered normal:
+This becomes manifest in a divergence of the functions $f_i$ as their argument
+approaches -1.  On the other hand, $f_1$, $f_2$, $f_3$ all go to zero as their
+argument approaches positive infinity, and hence for large $r$ the probability
+distribution of $\bar{\phi}$ is a zero-centered normal:
 \begin{equation}
@@
 obtain a system of equations\footnote{There are of course many alternative,
-arguably simpler, solutions that differ by subleading orders in an expansion in
+arguably simpler, solutions that differ by sub-leading orders in an expansion in
 $g(R_1)$. The one displayed here is the exact solution to
@@
     the shaded region shows where the perturbative expansion becomes invalid.
-    Similarly in the right plot, the lines become dotted oustside of the
+    Similarly in the right plot, the lines become dotted outside of the
     perturbative region. Note how the sign of $r$ is preserved, $r = 0$ is a
-    solution, and $g$ goes to $0$ for $R_2 \gg R_1$.}
+    solution, and $g$ vanishes for $R_2 \gg R_1$.}
 \end{figure}
@@
 
-Thus we conclude that, for sufficiently large system size: 
+Thus we conclude that, at the critical point, for sufficiently large system
+size:
 \begin{equation}
     \label{eq:log_pdf_critical_point}
-    -\log p(\phi) \sim \mathrm{const} + \frac{2\pi^2}{9} V g\bar{\phi}^4\,,
+    -\log p(\bar{\phi}) \sim \mathrm{const} + \frac{2\pi^2}{9} V g\bar{\phi}^4\,,
 \end{equation}
-with subleading terms vanishing like $1 / \log(R)$.
+with sub-leading terms vanishing like $1 / \log(R)$.
 
@@
 critical point\footnote{ Incidentally, we note that this result is not in
-contraddiction with the claim in the original paper \cite{Binder1981} that the
+contradiction with the claim in the original paper \cite{Binder1981} that the
 critical Binder cumulant in 4 dimensions is zero, because in the original paper
@@
 \section{Derivation on the 4-torus}
+\label{sect:derivation}
 
@@
     \phi(x) = \frac{1}{V} \sum_{n \in \mathbb{Z}^4} \phi_{n} 
-    e^{i \frac{n\cdot x}{R}}\,;\quad \bar{\phi} = \frac{\phi_{0}}{V}\,.
+    e^{i \frac{n\cdot x}{R}}\,;\quad \bar{\phi} = \frac{1}{V}\int\dd^4 x\ \phi(x) = \frac{1}{V}\phi_{n = 0}\,.
 \end{equation}
@@
 $\phi_{0}$ becomes an external field, whereas the other components $\phi_{n}$
-with are still part of the functional integral. Separating the zero momentum
-mode from the other modes in the action yields:
+are still part of the functional integral. Separating the zero momentum mode
+from the other modes in the action yields:
 \begin{equation}
@@
     V \left(\frac{1}{2} r_0 \bar{\phi}^2 + \frac{u_0}{24} \bar{\phi}^4\right) +
-    \frac{u_0}{4V} \bar{\phi}^2 \sum_{n} \phi_{n} \phi_{-n} + \\
+    \frac{1}{2V} \sum_{n} \left(\frac{n^2}{R^2} + r_0 +
+      \frac{1}{2} u_0\bar{\phi}^2 \right)\phi_{n} \phi_{-n} +\\
     &\frac{u_0}{6V^2} \bar{\phi} 
-        \sum_{n_1, n_2} \phi_{n_1} \phi_{n_2} \phi_{-n_1 -n_2} + \\
-    &\frac{u_0}{24V^3} \sum_{n_1, n_2, n_3} \phi_{n_1} \phi_{n_2} 
+        \sum_{n_1, n_2} \phi_{n_1} \phi_{n_2} \phi_{-n_1 -n_2} +
+    \frac{u_0}{24V^3} \sum_{n_1, n_2, n_3} \phi_{n_1} \phi_{n_2}
         \phi_{n_3} \phi_{-n_1 -n_2 -n_3}\,,
@@
 \end{equation}
-where all summations now exclude $n = (0, 0, 0, 0)$.
+where all summations now are over $\mathbb{Z}^4 \setminus \{0\}$.
 
@@
 \begin{tikzpicture}[scale=0.5, baseline=(current bounding box.center)]
+    \draw[thick] (-1, 0) -- (1, 0);
+    \node[left] at (-1, 0) {$n_1$};
+    \node[right] at (1, 0) {$n_2$};
+\end{tikzpicture}
+    \quad
+    \frac{V}{n_1^2R^{-2} + r}\delta_{n_1 + n_2}
+\quad\quad
+\begin{tikzpicture}[scale=0.5, baseline=(current bounding box.center)]
     \draw[densely dashed, thick] (-1, -1) -- (0, 0) -- (-1, 1);
@@
     -\frac{u\bar{\phi}^2}{4V}\delta_{n_1 + n_2}
-    \quad\quad
+\end{displaymath}
+\begin{displaymath}
 \begin{tikzpicture}[scale=0.5, baseline=(current bounding box.center)]
@@
     -\frac{u\bar{\phi}}{6V^2}\delta_{n_1 + n_2 + n_3}
-\end{displaymath}
-\begin{displaymath}
+\quad\quad
 \begin{tikzpicture}[scale=0.5, baseline=(current bounding box.center)]
@@
     -\frac{u}{24V^3}\delta_{n_1 + n_2 + n_3 + n_4}
-\quad\quad
-\begin{tikzpicture}[scale=0.5, baseline=(current bounding box.center)]
-    \draw[thick] (-1, 0) -- (1, 0);
-    \node[left] at (-1, 0) {$n_1$};
-    \node[right] at (1, 0) {$n_2$};
-\end{tikzpicture}
-    \quad
-    \frac{V}{n_1^2R^{-2} + r}\delta_{n_1 + n_2}
 \end{displaymath}
@@
     I_k = \frac{1}{V} \sum_{n\in\mathbb{Z}^4\setminus\{0\}} 
-    \frac{e^{-s(r + n^2 R^{-2})}}{\left(r + n^2 R^{-2}\right)^k} = 
-    \frac{1}{(k - 1)!}\left(-\frac{\partial}{\partial r} - s\right)^k I_1
+    \frac{e^{-s(n^2 R^{-2} + r)}}{\left(n^2 R^{-2} + r\right)^k} = 
+    \frac{1}{(k - 1)!}\left(-\frac{\partial}{\partial r} - s\right)^{k-1} I_1
 \end{equation}
@@
     I_1 &= \frac{1}{V} \int_{s}^{\infty} \dd t \sum_{n\in\mathbb{Z}^4\setminus\{0\}} 
-    e^{-t(r + n^2 R^{-2})}\\
+    e^{-t(n^2 R^{-2} + r)}\\
     & = \frac{1}{V} \int_{s}^{\infty} \dd t\ e^{-t r}
@@
     \label{eq:theta}
-    \theta(z) = \theta_3(0, e^{-z}) = \sum_{m = -\infty}^{\infty} e^{-m^2 z}\,.
+    \theta(z) = \theta_3(0; e^{-z}) = \sum_{m = -\infty}^{\infty} e^{-m^2 z}\,.
 \end{equation}
@@
 form. In order to do that, we need the asymptotic behavior of $\theta(z)$ for
-$z\to0$, which can be obtained from its definition using the Euler-Mclaurin
+$z\to0$, which can be obtained from its definition using the Euler-Maclaurin
 formula:
@@
 \begin{split}
-    -\log p(\bar{\phi}) =& \mathrm{const}+ V \Bigg( \\
-    &\frac{1}{2} \bar{\phi}^2\left( r + \frac{u}{32\pi^2}  \left(\left(\frac{1}{R^2} f(rR^2, \mu R) + r - \mu^2\right)\right) + O(u^2)\right) + \\
+    -\log p(\bar{\phi}) =& \mathrm{const}+ V \Bigg(
+    \frac{1}{2} \bar{\phi}^2\left( r + \frac{u}{32\pi^2R^2}  \left(f(rR^2, \mu R) + (r - \mu^2)R^2\right) + O(u^2)\right) + \\
     &\frac{u}{24} \bar{\phi}^4 \left(1 + \frac{3u}{32\pi^2} f^{(1, 0)}(rR^2, \mu R) + O(u^2)\right) + \\
@@
 \begin{equation}
-    S += \int \dd^4 x\ \frac{1}{2}\xi_0 \phi^2 \mathcal{R}\,,
+    \Delta S = \int_{\mathcal{M}} \frac{1}{2}\xi_0 \phi^2 \mathcal{R}\,,
 \end{equation}
 where $\mathcal{R}$ is the scalar curvature and $\xi_0$ is a dimensionless bare
-coupling. On a static manifold, this coupling is equivalent to a mass term.
-The free theory is Weyl invariant if $\xi_0 = \frac{1}{6}$. In the
-presence of interactions, the coupling needs to be renormalized as well, and
-its renormalized counterpart $\xi$ becomes a running coupling. At one loop, the
+coupling.  The free theory is Weyl invariant if $\xi_0 = \frac{1}{6}$. In the
+presence of interactions, the coupling needs to be renormalized, and its
+renormalized counterpart $\xi$ becomes a running coupling. At one loop, the
 renormalization group equation for $\xi$ is:
@@
 
-From this experssion it seems that $\xi$ can be set to the critical value
+From this expression it seems that $\xi$ can be set to the critical value
 $\frac{1}{6}$ at all energy scales. However, this turns out to be an illusion:
@@
 \begin{equation}
-    rR^2 \to \frac{\xi}{12} + r R^2\,
+    rR^2 \to \frac{\xi}{12} + r R^2\,,
 \end{equation}
@@
 \begin{align}
-    &f_1(\bar{r}) = (2 - \bar{r})\int \dd z\ e^{-z(\bar{r} + 4)} \left(H(z) - \frac{1}{z}\right) + \frac{7}{3}\,,\\
-    &f_2(\bar{r}) = \int \dd z\ e^{-z(\bar{r} + 4)} 
-    \left(1 + z (2 - \bar{r})\right) \left(H(z) - \frac{1}{z}\right)+ \frac{6}{\bar{r} + 4}\,,\\
-    &f_3(\bar{r}) = \int \dd z\ e^{-z(\bar{r} + 4)} 
-    \left(2 + z (2 - \bar{r})\right)z \left(H(z) - \frac{1}{z}\right) + \frac{\bar{r} + 10 }{(\bar{r} + 4)^2}\,,\\
+    &f_1(\bar{r}) = (2 - \bar{r})\int_{0}^{\infty} \dd z\ e^{-z(\bar{r} + 4)} H(z) + \frac{7}{3}\,,\\
+    &f_2(\bar{r}) = \int_{0}^{\infty} \dd z\ e^{-z(\bar{r} + 4)} 
+    \left(1 + z (2 - \bar{r})\right) H(z) + \frac{6}{\bar{r} + 4}\,,\\
+    &f_3(\bar{r}) = \int_{0}^{\infty} \dd z\ e^{-z(\bar{r} + 4)} 
+    \left(2 + z (2 - \bar{r})\right)z H(z) + \frac{\bar{r} + 10 }{(\bar{r} + 4)^2}\,,\\
 \end{align}
@@
 \begin{equation}
-    H(z) = \sum_{\ell = 1}^{\infty} (2\ell + 3)e^{-z (\ell(\ell + 3) - 4)}\,.
+    H(z) = \sum_{\ell = 1}^{\infty} (2\ell + 3)e^{-z (\ell(\ell + 3) - 4)} - \frac{1}{z}\,.
 \end{equation}
@@
 \end{document}
-
-The coupling $\eta$ is related to the coupling between the scalar and
-the curvature, and hence it is exactly zero on the tourus. On the sphere, it
-may seem that it can be set to a constant value $\eta = 2$ at all energy
-scales, however this ceases to be true at higher orders in perturbation theory,
-where the renormalization group equation becomes inhomogeneous.

### 2020-04-19 paper/paper.tex
Proofread

@@
 
-\title{Perturbative probability distribution of the average Ising magnetization
-in 4 dimensions.}
+\title{Perturbative PDF of the total magnetization of the 4D Ising model }
 
@@
 
-\abstract{We compute, at one loop in perturbation theory, the probability density
-function of the average magnetization $\bar{\phi}$ of the Ising model, on both
-the 4-torus and the 4-sphere. The perturbative expansion covers the entire
-symmetric phase and a neighborhood of the critical point. We find that, at the
-critical point, the PDF has the asymptotic behavior $p(\bar{\phi})\sim
-\exp(-c \bar{\phi}^4)$, and hence the critical value of the Binder cumulant
-is $U = 1 - \frac{4\,\Gamma(5/4)^2}{ 3\,\Gamma(3/4)^2}$. We validate our
-results by comparison with Monte Carlo simulation.}
+\abstract{We compute, at one loop in perturbation theory, the probability
+density function of the total magnetization $M$ of the Ising model, on both the
+4-torus and the 4-sphere. We develop a single perturbative expansion that is
+valid both in the symmetric phase, as long as the correlation length $\xi$ is
+large in lattice units, and in the broken symmetry phase, as long as the
+correlation length is large compared to the system size $L$. We find that, at
+the critical point, for very large system size in lattice units, the PDF
+approaches $p(M)\sim \exp(-f(L) M^4)$, and hence the critical value of the
+Binder cumulant of the total magnetization is $U = 1 -
+\frac{4\,\Gamma(5/4)^2}{3\,\Gamma(3/4)^2}$. We validate our results by
+comparison with Monte Carlo simulation.}
 \begin{document}
@@
 
-\section{Main results}
+\section{Main result}
 
-The average magnetization $\bar{\phi} = \frac{1}{V} \sum_{i} \sigma_i$ of the
-Ising model is the simplest observable in Monte Carlo simulations. When the
-system size $L$ is large, the probability distribution of $\bar{\phi}$ changes
-qualitatively between the two phases of the model. In the symmetric phase,
-\textit{i. e.} when the correlation length $\xi$ is large in lattice units, but
-small compared to the system size, the distribution is well approximated by a
-zero-mean normal. In the broken symmetry phase, still for $1 \ll \xi \ll L$,
-the distribution is well approximated by a mixture of two normals centered at
-non-zero values $\pm \bar{\phi}_0$.
+When the system size $L$ is large, the probability distribution of the total
+magnetization $M$ of the Ising model changes qualitatively between the two
+phases of the model. In the symmetric phase, \textit{i.e.} when the correlation
+length $\xi$ is large in lattice units, but small compared to the system size,
+the distribution is well approximated by a zero-mean normal.  In the broken
+symmetry phase, still for $1 \ll \xi \ll L$, the distribution is well
+approximated by a mixture of two normals centered at non-zero values $\pm
+M_0$.
 
-Near the critical point, \textit{ie.} $1 \ll L \ll \xi$, the probability
-distribution of $\bar{\phi}$ is a non-trivial function of $\bar{\phi}$, at
-least in two and three dimensions.  It is not immediately clear if the same is
-also true in four dimensions. Since the field theory that describes the Ising
-critical point in four dimensions becomes weakly coupled at low energy, one may
-expect the probability distribution of $\bar{\phi}$ to remain a zero-centered
-normal at the critical point as well as in the symmetric phase.
-
-To clarify this issue, we compute the logarithm of the PDF of $\bar{\phi}$ at
-one-loop in $4$ dimensions, and show that, for very large system size, the
-critical probability distribution is not normal, but rather of the form $-\log
-p(\bar{\phi})\sim f(L) \bar{\phi}^4$. We find that corrections to this form
-vanish very slowly with increasing system size, like $1/\log L$.
+Near the critical point $1 \ll L \ll \xi$, in two and three dimensions, the
+probability distribution of $M$ is a non-trivial function of $M$.  It is not
+immediately clear if the same is also true in four dimensions. Since the field
+theory that describes the Ising critical point in four dimensions becomes
+weakly coupled at low energy, one may expect the probability distribution of
+$M$ to remain a zero-centered normal even at the critical point. This is in
+fact the case for the total magnetization of a subsystem of intermediate size:
+large in lattice units, but small compared to $L$ \cite{Binder1981}. Here we
+show that the same is not true for the total magnetization of the entire
+system. We compute the logarithm of the PDF of $M$ at one-loop in $4$
+dimensions, and show that, for very large system size, the critical probability
+distribution is not normal, but rather of the form $-\log p(M)\sim f(L) M^4$.
+We find that corrections to this form vanish very slowly with increasing system
+size, like $1/\log L$.
 
@@
 
-Thus we consider the $\phi^4$ field theory with action:
+When the correlation length is large in lattice unit, the Ising model in 4
+dimensions is described by the $\phi^4$ field theory:
 \begin{equation}
-  S[\phi] \equiv \int \dd^4 x\ \left(
-    \frac{1}{2} \phi(x)\left(\Delta + r_0\right)\phi(x) +
-    \frac{1}{24} u_0 \phi(x)^4\right)
+    S[\phi] \equiv \int_{\mathcal{M}} \left(
+    \frac{1}{2} \phi\left(\Delta + r_0\right)\phi +
+    \frac{1}{24} u_0 \phi^4\right)\,.
 \end{equation}
-defined on a 4-torus of radius $R$, i.e. $x_i\sim x_i + 2\pi R$. In this field
-theory context, the PDF of the average magnetization is defined as:
+
+Here we take the manifold $\mathcal{M}$ to be a 4-torus of radius $R$,
+\textit{i.e.} $x_i\sim x_i + 2\pi R$. In this field theory context, we define
+the PDF of the average magnetization $m = M / V$ as:
 \begin{equation}
     \label{eq:pdf_definition}
-    p(\bar\phi) \equiv \frac{1}{Z} \int \left[\mathcal{D}\phi\right]
-    \delta\left(\bar{\phi} - \frac{1}{V}\int\dd^4 x\ \phi(x)\right)
+    p(m) \equiv \frac{1}{Z} \int \left[\mathcal{D}\phi\right]
+    \delta\left(m - \frac{1}{V}\int\dd^4 x\ \phi(x)\right)
     e^{-S[\phi]}\,,
 \end{equation}
-and at one loop it evaluates to:
+and we evaluate it at one loop in perturbation theory, obtaining:
 \begin{equation}
@@
 \begin{split}
-    -\log p(\bar{\phi}) = \mathrm{const.} +  V\Bigg(
-  &\frac{1}{2} \frac{\bar{\phi}^2}{R^2} \left(
+    -\log p(m) = \mathrm{const.} +  V\Bigg(
+  &\frac{1}{2} \frac{m^2}{R^2} \left(
     r R^2 + \frac{g}{6} \left(f_1\left(r R^2\right) - 1\right)
     + O\left(g^2\right)\right) + \\
-  & \frac{2\pi^2}{9}g\bar{\phi}^4\left(
+  & \frac{2\pi^2}{9}gm^4\left(
     1 - \frac{g}{2}f_2\left(r R^2\right)
     + O\left(g^2\right)\right) + \\
-  & \frac{8\pi^4}{81} g^3 \bar{\phi}^6 R^2\left(
+  & \frac{8\pi^4}{81} g^3 m^6 R^2\left(
     f_3\left(r R^2\right)
-    + O\left(g\right)\right) + O\left(g^4\bar{\phi}^8\right) \Bigg)\,,
+    + O\left(g\right)\right) + O\left(g^4m^8\right) \Bigg)\,.
 \end{split}
 \end{equation}
-where the coefficients $r$ and $g$ are the renormalized counterparts to $r_0$
-and $u_0$. An explicit expression for the functions $f_1$, $f_2$ and $f_3$ is
-reported in section \ref{sect:derivation} (\ref{eq:f_functions}), and they are
-plotted in
-fig.~\ref{fig:f_functions}.
+Here the coefficients $r$ and $g$ are the renormalized counterparts to $r_0$
+and $u_0$; the functions $f_1$, $f_2$ and $f_3$ are plotted in
+fig.~\ref{fig:f_functions}, and an explicit expression (\ref{eq:f_functions})
+is given below.
 
@@
 
-Eq. (\ref{eq:log_pdf}) is our central result, and we now describe it
-at length. We first illustrate its domain of validity and its dependence on the
-renormalized couplings $r$, $g$ on a torus of fixed radius $R$, then describe
-its dependence on $R$ at fixed bare couplings $r_0$, $u_0$.
+This is our central result, and we now discuss it at length. We first
+illustrate its domain of validity and its dependence on the renormalized
+couplings $r$, $g$ on a torus of fixed radius $R$, then describe its dependence
+on $R$ at fixed bare couplings $r_0$, $u_0$.
 
 The dimension-2 coupling $r$ controls the cross-over between the symmetric and
-broken-symmetry phases. The expression (\ref{eq:log_pdf}) is valid for all
-positive values of $r$, and also for negative values, provided that $rR^2
-\gtrsim -1/2$.  For lower values of $r$, additional modes of the Laplacian
-condense beside the zero-mode, and this invalidates the perturbative expansion.
-This becomes manifest in a divergence of the functions $f_i$ as their argument
-approaches -1.  On the other hand, $f_1$, $f_2$, $f_3$ all go to zero as their
-argument approaches positive infinity, and hence for large $r$ the probability
-distribution of $\bar{\phi}$ is a zero-centered normal:
+broken-symmetry phases. The expression (\ref{eq:log_pdf}) is valid for  $rR^2
+\gtrsim -1/2$.  For lower values of $r$, the perturbative vacuum becomes
+unstable, invalidating the perturbative expansion. For this reason, the
+functions $f_i$ diverge as their argument approaches -1.  On the other hand,
+$f_1$, $f_2$, $f_3$ all go to zero as their argument approaches positive
+infinity, and hence for large $r$ the probability distribution of $m$ is a
+zero-centered normal:
 \begin{equation}
-    p(\bar{\phi}) \sim N e^{- \frac{1}{2} V r \bar{\phi}^2}
+    p(m) \sim N e^{- \frac{1}{2} V r m^2}
   \quad \text{for} \quad
@@
 The dimensionless coupling $g$ is the parameter of the perturbative expansion.
-The expansion is valid for $g \ll 1$, provided $rR^2$ is sufficiently far from
-the bound discussed above.
+The expansion is valid for $g \lesssim 1$, provided $rR^2$ is sufficiently far
+from the bound discussed above.
 
-In order to describe how $p(\bar{\phi})$ depends on $R$ at fixed bare couplings
-$r_0$, $u_0$, it is necessary to account for renormalization effects. In
-general, the renormalized couplings $r$, $g$ depend on a renormalization scale
-$\mu$, and it is possible to derive an expression for $p(\bar{\phi})$ evaluated
-at a generic value of $\mu$. However, for simplicity,
-(\ref{eq:log_pdf}) is evaluated at the scale $\mu^2 = r + R^{-2}$.
-This choice of $\mu$ is optimal for the reliability of perturbation theory,
-because it avoids the emergence of large logarithms over the widest possible
-range of parameters.
+In order to describe how $p(m)$ depends on $R$ at fixed bare couplings $r_0$,
+$u_0$, it is necessary to account for renormalization effects. The specific
+renormalization scheme we used is described in section \ref{sect:derivation},
+and (\ref{eq:log_pdf_with_mu}) gives expression for $p(m)$ evaluated at a
+generic renormalization scale $\mu$. However, for simplicity, we chose to
+evaluate (\ref{eq:log_pdf}) at the scale $\mu^2 = r + R^{-2}$.  This choice of
+$\mu$ is optimal for the reliability of perturbation theory, because it avoids
+the emergence of large logarithms over the widest possible range of parameters.
 
-The renormalization group equations for $r$ and $g$ are:
+At one loop, the Callan-Symanzik equations for $r$ and $g$ are:
 \begin{equation}
-  \label{eq:renormalization_group}
+  \label{eq:callan_symanzik}
   \mu \frac{\dd g}{\dd \mu} = g^2\,,\quad
@@
 $g(R_1)$. The one displayed here is the exact solution to
-(\ref{eq:renormalization_group}).} connecting the renormalized couplings at two
+(\ref{eq:callan_symanzik}).} connecting the renormalized couplings at two
 different values of $R$:
 \begin{equation}
+  \label{eq:renormalized_relation}
   \frac{g(R_1)}{g(R_2)} = \frac{r(R_1)^3}{r(R_2)^3} = 
@@
     \caption{\label{fig:renormalization_group} A few solutions to
-    (\ref{eq:renormalization_group}) with $g(R_1) = 0.45$.  In the left plot,
-    the shaded region shows where the perturbative expansion becomes invalid.
-    Similarly in the right plot, the lines become dotted outside of the
-    perturbative region. Note how the sign of $r$ is preserved, $r = 0$ is a
-    solution, and $g$ vanishes for $R_2 \gg R_1$.}
+    (\ref{eq:renormalized_relation}) with $g(R_1) = 0.45$.  In the left plot,
+    the shaded region shows where the perturbative vacuum becomes unstable,
+    invalidating the perturbative expansion. Similarly, in the right plot, the
+    lines become dotted outside of the perturbative region. Note how the sign
+    of $r$ is preserved and $r = 0$ is a solution.}
 \end{figure}
 
-A few solutions to these equations are shown in
-fig.~\ref{fig:renormalization_group}. Within the perturbative regime $g \ll 1$,
-$r R^2 \gtrsim -1/2$, the coupling $r$ varies very little with $R$, and, as is
-clear from the differential form (\ref{eq:renormalization_group}), the sign of
-$r$ is always preserved. Thus we conclude that the critical point is at $r
-= 0$, and the symmetric phase is realized for $r > 0$.
+A few solutions to this system of equations are shown in
+fig.~\ref{fig:renormalization_group}. Within the perturbative regime $g
+\lesssim 1$, $r R^2 \gtrsim -1/2$, the coupling $r$ varies little with
+$R$, and, as is clear from the differential form (\ref{eq:callan_symanzik}),
+the sign of $r$ is always preserved. Thus we conclude that the critical point
+is at $r = 0$, and the symmetric phase is realized for $r > 0$.
 
-At the critical point, the renormalized coupling $g$ follows the simplified renormalization group equation:
+At the critical point, the renormalized coupling $g$ follows the simplified
+Callan-Symanzik equation
 \begin{equation}
@@
 \end{equation}
-and hence, as system size grows, the renormalized coupling $g$ goes to zero.
-In this regime, the quartic term in (\ref{eq:log_pdf}) dominates all the
-others. This is perhaps most evident if the PDF is expressed in terms of the
-rescaled quantity $\bar{\varphi} = g^{\frac{1}{4}} \bar{\phi}$, whose
-variance remains finite as $g\to0$.
-
-Thus we conclude that, at the critical point, for sufficiently large system
-size:
+and hence, as system size grows, the renormalized coupling $g$ goes to zero as
+$1 / \log(R)$.  In this regime, the quartic term in (\ref{eq:log_pdf})
+dominates all the others. This is perhaps most evident if the PDF is expressed
+in terms of the rescaled quantity $\bar{m} = g^{\frac{1}{4}} m$, whose variance
+remains finite as $g\to0$. Thus we conclude that, at the critical point, for
+sufficiently large system size:
 \begin{equation}
     \label{eq:log_pdf_critical_point}
-    -\log p(\bar{\phi}) \sim \mathrm{const} + \frac{2\pi^2}{9} V g\bar{\phi}^4\,,
+    -\log p(m) \sim \mathrm{const} + \frac{2\pi^2}{9} V gm^4\,.
 \end{equation}
-with sub-leading terms vanishing like $1 / \log(R)$.
 
@@
 \begin{equation}
-    U = 1 - \frac{\left\langle \bar \phi^4 \right \rangle}
-    {3\left\langle \bar \phi^2 \right \rangle^2},
+    \label{eq:binder_cumulant}
+    U = 1 - \frac{\left\langle m^4 \right \rangle}
+    {3\left\langle m^2 \right \rangle^2}\,.
 \end{equation}
-which is constructed to be independent of the overall scale of $\bar{\phi}$,
-and to be zero if $\bar{\phi}$ is normally distributed. From
+
+This quantity is constructed to be independent of the overall scale of $m$,
+and to be zero if $m$ is normally distributed. From
 (\ref{eq:log_pdf_critical_point}) we conclude that, on a 4-Torus, at the
-critical point\footnote{ Incidentally, we note that this result is not in
-contradiction with the claim in the original paper \cite{Binder1981} that the
-critical Binder cumulant in 4 dimensions is zero, because in the original paper
-the cumulant is not computed from the total magnetization, but rather from the
-average magnetization of blocks of spins much smaller than the overall system
-size.}:
+critical point:
 \begin{equation}
@@
 computed from (\ref{eq:log_pdf}) near the critical point, and the results of
-Monte Carlo simulation of the $4D$ Ising model. The agreement is excellent except
-for the smallest system size $L = 8$, probably because of contributions from
-irrelevant operators. Notice how slowly the finite size Binder cumulant
-approaches the asymptotic value (\ref{eq:critical_binder_cumulant}).
+Monte Carlo simulation of the $4D$ Ising model. The agreement is excellent
+except for the smallest system size $L = 8$. Notice how slowly the finite size
+Binder cumulant approaches the asymptotic value
+(\ref{eq:critical_binder_cumulant}).
 
@@
 We now describe briefly how the result (\ref{eq:log_pdf}) is obtained. The
-perturbative approach is very similar to the computation of the effective
+perturbative approach is similar to the computation of the effective
 action, as in \textit{e.g.} \cite{WeinbergEffectiveAction}, except that we are
-interested in the whole probability distribution of $\bar{\phi}$, instead of
+interested in the whole probability distribution of $m$, instead of
 just the expected value. The main difficulty lies in evaluating the loop
@@
 
-First of all we observe that the average magnetization
-$\bar{\phi}$ is proportional to the zero momentum mode of the field $\phi$:
+The average magnetization $m$ is proportional to the zero-momentum mode of the
+field:
 \begin{equation}
     \phi(x) = \frac{1}{V} \sum_{n \in \mathbb{Z}^4} \phi_{n} 
-    e^{i \frac{n\cdot x}{R}}\,;\quad \bar{\phi} = \frac{1}{V}\int\dd^4 x\ \phi(x) = \frac{1}{V}\phi_{n = 0}\,.
+    e^{i \frac{n\cdot x}{R}}\,;\quad m = \frac{1}{V}\int\dd^4 x\ \phi(x) = \frac{1}{V}\phi_{n = 0}\,.
 \end{equation}
 
-Thus, because of the delta function in (\ref{eq:pdf_definition}), the component
-$\phi_{0}$ becomes an external field, whereas the other components $\phi_{n}$
-are still part of the functional integral. Separating the zero momentum mode
-from the other modes in the action yields:
+Because of the delta function in (\ref{eq:pdf_definition}), the zero-mode
+becomes an external field, whereas all other modes are still part of the
+functional integral. Separating the zero-mode from the other modes in the
+action yields:
 \begin{equation}
 \begin{split}
-    S(\bar{\phi}, \phi) =& 
-    V \left(\frac{1}{2} r_0 \bar{\phi}^2 + \frac{u_0}{24} \bar{\phi}^4\right) +
+    S(m, \phi) =& 
+    V \left(\frac{1}{2} r_0 m^2 + \frac{u_0}{24} m^4\right) +
     \frac{1}{2V} \sum_{n} \left(\frac{n^2}{R^2} + r_0 +
-      \frac{1}{2} u_0\bar{\phi}^2 \right)\phi_{n} \phi_{-n} +\\
-    &\frac{u_0}{6V^2} \bar{\phi} 
+      \frac{1}{2} u_0m^2 \right)\phi_{n} \phi_{-n} +\\
+    &\frac{u_0}{6V^2} m 
         \sum_{n_1, n_2} \phi_{n_1} \phi_{n_2} \phi_{-n_1 -n_2} +
@@
 
-We also introduce renormalized couplings
+We introduce renormalized couplings
 \begin{equation}
@@
 \end{tikzpicture}
-    -\frac{u\bar{\phi}^2}{4V}\delta_{n_1 + n_2}
+    -\frac{um^2}{4V}\delta_{n_1 + n_2}
 \end{displaymath}
@@
 \end{tikzpicture}
-    -\frac{u\bar{\phi}}{6V^2}\delta_{n_1 + n_2 + n_3}
+    -\frac{um}{6V^2}\delta_{n_1 + n_2 + n_3}
 \quad\quad
@@
 
-The logarithm of the probability distribution of $\bar{\phi}$ is the sum of all
+The logarithm of the probability distribution of $m$ is the sum of all
 connected diagrams:
@@
 \begin{split}
-    \log p(\bar{\phi}) = &\,
-    \mathrm{const} -V \left(\frac{1}{2} r \bar{\phi}^2 + 
-    \frac{u}{24} \bar{\phi}^4\right)\\
+    \log p(m) = &\,
+    \mathrm{const} -V \left(\frac{1}{2} r m^2 + 
+    \frac{u}{24} m^4\right)\\
 &+\ 
@@
 \begin{split}
-    -\log p(\bar{\phi}) = \mathrm{const} + V \Bigg( 
-    &\frac{1}{2} \bar{\phi}^2\left( r + u \left(\delta r + \frac{1}{2} I_1\right) + O(u^2)\right) + \\
-    &\frac{u}{24} \bar{\phi}^4 \left(1 + u \left(\delta u - \frac{3}{2} I_2\right) + O(u^2)\right) + \\
-    &\frac{u^3}{48}\bar{\phi}^6\left(I_3 + O(u)\right) + O(u^4 \bar{\phi}^8) \Bigg)\,,
+    -\log p(m) = \mathrm{const} + V \Bigg( 
+    &\frac{1}{2} m^2\left( r + u \left(\delta r + \frac{1}{2} I_1\right) + O(u^2)\right) + \\
+    &\frac{u}{24} m^4 \left(1 + u \left(\delta u - \frac{3}{2} I_2\right) + O(u^2)\right) + \\
+    &\frac{u^3}{48}m^6\left(I_3 + O(u)\right) + O(u^4 m^8) \Bigg)\,,
 \end{split}
@@
     &\delta r = -\frac{1}{32\pi^2}\left(\frac{1}{s} + r\left(\log s \mu^2 + \gamma_E - 1\right)\right)\,,\\
-    &\delta u = -\frac{3}{32\pi^2}\left(\log s\mu^2 +\gamma_E + 1\right)\,,
+    &\delta u = -\frac{3}{32\pi^2}\left(\log s\mu^2 +\gamma_E + 1\right)\,.
 \end{align}
-from which the renormalization group equations (\ref{eq:renormalization_group})
-can be obtained, and we have:
+The Callan-Symanzik equations (\ref{eq:callan_symanzik}) follow from this
+subtraction choice.
+
+Finally, we obtain:
 \begin{equation}
 \begin{split}
-    -\log p(\bar{\phi}) =& \mathrm{const}+ V \Bigg(
-    \frac{1}{2} \bar{\phi}^2\left( r + \frac{u}{32\pi^2R^2}  \left(f(rR^2, \mu R) + (r - \mu^2)R^2\right) + O(u^2)\right) + \\
-    &\frac{u}{24} \bar{\phi}^4 \left(1 + \frac{3u}{32\pi^2} f^{(1, 0)}(rR^2, \mu R) + O(u^2)\right) + \\
-    &\frac{u^3}{48}\bar{\phi}^6\left(\frac{R^2}{32\pi^2}f^{(2, 0)}(rR^2, \mu R) + O(u)\right) + O(u^4 \bar{\phi}^8) \Bigg)\,,
+    \label{eq:log_pdf_with_mu}
+    -\log p(m) =& \mathrm{const}+ V \Bigg(
+    \frac{1}{2} m^2\left( r + \frac{u}{32\pi^2R^2}  \left(f(rR^2, \mu R) + (r - \mu^2)R^2\right) + O(u^2)\right) + \\
+    &\frac{u}{24} m^4 \left(1 + \frac{3u}{32\pi^2} f^{(1, 0)}(rR^2, \mu R) + O(u^2)\right) + \\
+    &\frac{u^3}{48}m^6\left(\frac{R^2}{32\pi^2}f^{(2, 0)}(rR^2, \mu R) + O(u)\right) + O(u^4 m^8) \Bigg)\,,
 \end{split}
@@
 \begin{equation}
-    \int_0^{\infty} \dd z\, \frac{e^{-a z} - e^{-b z}}{z} = \log \frac{b}{a}\,;
+    \int_0^{\infty} \dd z\, \frac{e^{-a z} - e^{-b z}}{z} = \log \frac{b}{a}\,.
 \end{equation}
-when the asymptotic behavior of the two terms for large $z$ is not well
+When the large-$z$ asymptotic behavior of the two terms is not well
 matched, the integral becomes large in magnitude, making the perturbative
@@
 \label{sec:four_sphere}
-It is possible to obtain $\log p(\bar{\phi})$ on the 4 sphere as well, with
-very similar methods. Here we highlight the main differences from the 4 torus.
+It is possible to obtain $p(m)$ on the 4 sphere as well, with
+similar methods. Here we highlight the main differences from the 4 torus.
 
@@
 renormalized counterpart $\xi$ becomes a running coupling. At one loop, the
-renormalization group equation for $\xi$ is:
+Callan-Symanzik equation for $\xi$ is:
 \begin{equation}
@@
 $\frac{1}{6}$ at all energy scales. However, this turns out to be an illusion:
-at higher orders in perturbation theory the renormalization group equation
+at higher orders in perturbation theory the Callan-Symanzik equation
 becomes inhomogeneous \cite{Brown1980}. Therefore, on a curved manifold, $\xi$

### 2020-04-28 paper/paper.tex
Proofread

@@
 \affiliation[a]{Independent scholar}
-\emailAdd{allais.andrea@gmail.com}
+\emailAdd{a\_allais@alum.mit.edu}
 
 \abstract{We compute, at one loop in perturbation theory, the probability
-density function of the total magnetization $M$ of the Ising model, on both the
+density function of the total magnetization $M$ of the Ising model on the
 4-torus and the 4-sphere. We develop a single perturbative expansion that is
@@
 large in lattice units, and in the broken symmetry phase, as long as the
-correlation length is large compared to the system size $L$. We find that, at
-the critical point, for very large system size in lattice units, the PDF
-approaches $p(M)\sim \exp(-f(L) M^4)$, and hence the critical value of the
-Binder cumulant of the total magnetization is $U = 1 -
+correlation length is large both in lattice units and compared to the system
+size $L$. We find that, at the critical point, for large system size in lattice
+units, the PDF approaches $p(M)\sim \exp(-f(L) M^4)$. Consequently, the
+critical value of the Binder cumulant of the total magnetization is $U = 1 -
 \frac{4\,\Gamma(5/4)^2}{3\,\Gamma(3/4)^2}$. We validate our results by
@@
 magnetization $M$ of the Ising model changes qualitatively between the two
-phases of the model. In the symmetric phase, \textit{i.e.} when the correlation
+phases of the model. In the symmetric phase, if the correlation
 length $\xi$ is large in lattice units, but small compared to the system size,
@@
 
-Near the critical point $1 \ll L \ll \xi$, in two and three dimensions, the
+In the opposite regime, $1 \ll L \ll \xi$, in two and three dimensions, the
 probability distribution of $M$ is a non-trivial function of $M$.  It is not
@@
 weakly coupled at low energy, one may expect the probability distribution of
-$M$ to remain a zero-centered normal even at the critical point. This is in
-fact the case for the total magnetization of a subsystem of intermediate size:
-large in lattice units, but small compared to $L$ \cite{Binder1981}. Here we
-show that the same is not true for the total magnetization of the entire
-system. We compute the logarithm of the PDF of $M$ at one-loop in $4$
-dimensions, and show that, for very large system size, the critical probability
-distribution is not normal, but rather of the form $-\log p(M)\sim f(L) M^4$.
-We find that corrections to this form vanish very slowly with increasing system
-size, like $1/\log L$.
+$M$ to remain a zero-centered normal even in this regime. This is in fact the
+case for the total magnetization of a subsystem of intermediate size: large in
+lattice units, but small compared to $L$ \cite{Binder1981}. Here we show that
+the same is not true for the total magnetization of the entire system. We
+compute the logarithm of the PDF of $M$ at one-loop in $4$ dimensions, and show
+that, for $1 \ll L \ll \xi$, the distribution is not normal, but rather of the
+form $-\log p(M)\sim f(L) M^4$.  We find that corrections to this form vanish
+very slowly with increasing system size, like $1/\log L$.
 
@@
     p(m) \equiv \frac{1}{Z} \int \left[\mathcal{D}\phi\right]
-    \delta\left(m - \frac{1}{V}\int\dd^4 x\ \phi(x)\right)
+    \delta\left(m - \frac{1}{V}\int_{\mathcal{M}} \phi\right)
     e^{-S[\phi]}\,,
@@
 
-This is our central result, and we now discuss it at length. We first
-illustrate its domain of validity and its dependence on the renormalized
-couplings $r$, $g$ on a torus of fixed radius $R$, then describe its dependence
-on $R$ at fixed bare couplings $r_0$, $u_0$.
+%This is our central result, and we now discuss it at length. We first
+%illustrate its domain of validity and its dependence on the renormalized
+%couplings $r$, $g$ on a torus of fixed radius $R$, then describe its dependence
+%on $R$ at fixed bare couplings $r_0$, $u_0$.
 
@@
 
-At the critical point, the renormalized coupling $g$ follows the simplified
-Callan-Symanzik equation
+At the critical point, the renormalized coupling $g$ follows the simpler
+Callan-Symanzik equation:
 \begin{equation}
@@
 
-Retaining only one-loop diagrams, and employing a heath kernel regulator we have:
+Retaining only one-loop diagrams, and employing a heat kernel regulator we have:
 \begin{equation}
